---
title: "01_Acropora"
author: "Magena Marzonie and Monique White"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Load Packages
library(dplyr)
library(vcfR)
library(adegenet)
library(data.table)
```

**This markdown is for cleaning genetic data, importing metadata, merging files, and filtering loci**
 # Section 1. Data Wrangle

#Load data

```{r}
vcf <- read.vcfR("CAGRF24080074-2_variants_shortened.vcf")

# Clean sample names
sample_names <- colnames(vcf@gt)[-1]
shortened_names <- sub("^([^-]+-[^-]+).*", "\\1", sample_names)
colnames(vcf@gt)[-1] <- shortened_names
```

#1. QC Filtering - all samples 
##1.1 Convert genotypes to numeric format

```{r}
gt <- extract.gt(vcf, element = "GT", as.numeric = FALSE)
geno_num <- matrix(NA, nrow = nrow(gt), ncol = ncol(gt))

for (i in 1:nrow(gt)) {
  for (j in 1:ncol(gt)) {
    g <- gt[i, j]
    geno_num[i, j] <- ifelse(g %in% c("0/0", "0|0"), 0,
                       ifelse(g %in% c("0/1", "1/0", "0|1", "1|0"), 1,
                       ifelse(g %in% c("1/1", "1|1"), 2, NA)))
  }
}
```

##1.2 Loci filtering

```{r}
## 1.2 Loci filtering

# Locus missingness
missing_per_locus <- rowMeans(is.na(geno_num))

# Minor allele count (MAC)
# Calculate MAF per locus
maf <- apply(geno_num, 1, function(x) {
  x <- x[!is.na(x)]
  p <- mean(x) / 2         # allele frequency of ALT allele
  min(p, 1 - p)            # minor allele frequency
})

# Keep loci: missingness ≤ 20%, MAF ≥ 0.05
keep_loci <- which(missing_per_locus <= 0.2 & maf >= 0.05)

# Apply locus filter
vcf_filtered <- vcf[keep_loci, ]

# Individual filtering
gt_filtered <- extract.gt(vcf_filtered, element = "GT", as.numeric = FALSE)
missing_per_ind <- colMeans(is.na(gt_filtered))
keep_inds <- names(missing_per_ind[missing_per_ind <= 0.5])
vcf_filtered <- vcf_filtered[, c(TRUE, colnames(gt_filtered) %in% keep_inds)]
saveRDS(missing_per_ind, file = "missing_per_ind.RData")

# Remove 2020 individuals
keep_inds <- colnames(vcf_filtered)[-1]
vcf_filtered <- vcf_filtered[, c(TRUE, !grepl("-20", keep_inds))]
```

##1.3 Genind conversion / Joining metadata

```{r}
acro_gen_full <- vcfR2genind(vcf_filtered)

# Read metadata
meta <- fread("merged_metadata.csv")

# Check for match
inds <- indNames(acro_gen_full)
inds_clean <- trimws(inds)
meta_clean <- meta %>% mutate(SampleID = trimws(as.character(SampleID)))

write.csv(x = meta_clean, file ="meta_clean.csv")

# Filter out '-20' samples from inds
inds_filtered <- inds_clean[!grepl("-20|-24b", inds_clean)]


# Filter metadata to match and remove '-20' individuals
meta_ordered <- meta_clean %>%
  filter(SampleID %in% inds_filtered) %>%
  arrange(match(SampleID, inds_filtered))
write.csv(meta_ordered, file = "meta_ordered.csv")

# Also update genind object to remove those individuals
acro_gen_full <- acro_gen_full[!grepl("-20", indNames(acro_gen_full)), ]
```

Step 6: Read and merge metadata

```{r}
library(data.table)
# Read metadata file and adjust SampleID column for merging
acro_meta <- read.csv("Acro_Meta2024.csv") %>%
  rename(SampleID = VialID)

# Match and merge metadata
inds <- indNames(acro_gen_full)
meta_trimmed <- merge(data.table(SampleID = inds), acro_meta, by = "SampleID", all.x = TRUE)
meta_trimmed <- meta_trimmed[match(inds, meta_trimmed$SampleID), ]

write.csv(x = meta_trimmed, file = "meta_trimmed.csv")

# Add population info (STRATA) to genind
acro_gen_full@pop <- as.factor(meta_trimmed$STRATA)

# Optionally, read and reassign population map if needed
pop_map <- read.table("strata.tsv", header=TRUE, stringsAsFactors = TRUE)
acro_gen_full@pop <- pop_map$STRATA
```

```{r}
pop_vector <- meta_trimmed$Reef  
pop(acro_gen_full) <- as.factor(pop_vector)

acro_gen_full <- acro_gen_full[pop(acro_gen_full) != "GBR", ]

```

```{r}
save(acro_gen_full, file = "acro_gen_full.RData")

load("acro_gen_full.RData")
```

```{r}
coords <- meta_trimmed[match(indNames(acro_gen_full), meta_trimmed$SampleID), c("GPS.S", "GPS.E")]
```

```{r}
write.table(coords,
            file = "coords.txt",
            sep = "\t",
            quote = FALSE,
            row.names = TRUE,
            col.names = FALSE)
```



```{r}
# Read raw lines
lines <- readLines("geno_matrix.txt")
length(lines)      # number of rows

# Check the number of columns per line
col_counts <- sapply(lines, function(x) length(strsplit(x, "\\s+")[[1]]))
table(col_counts)  # see the distribution of column counts
```

# Section 2. PCA
```{r}
library(dplyr)
library(vcfR)
library(adegenet)
library(data.table)
library(readr)
read.csv("meta_clean.csv")
```

##1.4 PCA of genetic variance
```{r}
load("acro_gen_full.Rdata")

x <- tab(acro_gen_full, NA.method = "mean")
pca1 <- dudi.pca(x, scannf = FALSE, scale = FALSE, nf = 3)

percent <- pca1$eig / sum(pca1$eig) * 100
barplot(percent, ylab = "Variance explained (%)", names.arg = round(percent, 1))
```
##1.5 Prepare PCA coordinates + add population info
```{r}
# Extract PCA coordinates
ind_coords <- as.data.frame(pca1$li)
# Add individual names
ind_coords$ind <- rownames(ind_coords)
ind_coords <- ind_coords %>% rename(SampleID = ind)
meta_clean <- read.csv("meta_clean.csv")

ind_coords <- ind_coords %>% left_join(meta_clean, by = "SampleID")

save(ind_coords, file = "ind_coords.RData")
```

##1.6 Plot PCA by reef
```{r}
library(ggplot2)

reef_order <- c(
  "Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
  "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
  "Wreck", "Cato", "GBR", "Elizabeth", "Middleton"
)

ind_coords$Reef <- factor(ind_coords$Reef, levels = reef_order)

ggplot(ind_coords, aes(x = Axis1, y = Axis2, color = Reef)) +
  geom_point(size = 3, alpha = 0.9) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "Population"
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank())

```
##1.7 Plot PCA by latitude
```{r}
library(wesanderson)
library(ggplot2)
library(wesanderson)

# Pick your fave Wes Anderson palette (e.g., "Darjeeling1", "Zissou1", "Moonrise1", etc.)
palette <- wes_palette("Zissou1", n = 100, type = "continuous")

ggplot(ind_coords, aes(x = Axis1, y = Axis2, color = GPS.S)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_gradientn(colors = palette) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "Latitude"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

```

##1.8 Plot PCA by longitude
```{r}
library(wesanderson)
library(ggplot2)
library(wesanderson)

# Pick your fave Wes Anderson palette (e.g., "Darjeeling1", "Zissou1", "Moonrise1", etc.)
palette <- wes_palette("Cavalcanti1", n = 100, type = "continuous")

ggplot(ind_coords, aes(x = Axis1, y = Axis2, color = GPS.E)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_gradientn(colors = palette) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "Longitude"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

```
##1.9 Plot PCA by depth
```{r}
library(wesanderson)
library(ggplot2)
library(wesanderson)

# Pick your fave Wes Anderson palette (e.g., "Darjeeling1", "Zissou1", "Moonrise1", etc.)
palette <- wes_palette("GrandBudapest1", n = 100, type = "continuous")

ggplot(ind_coords, aes(x = Axis1, y = Axis2, color = Depth..m.)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_gradientn(colors = palette) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "Depth (m)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )
```

# PCA with WGS samples highlighted

```{r}
wgs_samples <- c(
"A025-24", "A035-24", "A235-24","A052-24", "A196-24", "A066-24", "A214-24", "A334-24", "A342-24", "A353-24", "A100-24", "A115-24", "A277-24", "A290-24",
"I015-24", "I034-24", "I066-24", "I205-24", "I209-24", "I113-24", "I095-24", "I171-24", "I214-24", "I053-24", "I108-24", "I126-24",
"M124-24", "M132-24", "M229-24", "M047-24", "M228-24", "M128-24", "M027-24", "M069-24", "M102-24", "M107-24", "M111-24", "M224-24", "M186-24","M247-24", "M263-24",
"PD046-24", "PD036-24", "PD141-24", "PV119-24", "PV167-24","PV286-24", "PV314-24", "PM124-24", "PM384-24", "PM471-24"
)

library(dplyr)

ind_coords <- ind_coords %>%
  mutate(WGS = if_else(SampleID %in% wgs_samples, "yes", "no")) %>% 
  mutate(WGS = as.factor(WGS))
```

```{r}
ggplot(ind_coords, aes(x = Axis1, y = Axis2)) +
  geom_point(
    data = subset(ind_coords, WGS == "no"),
    aes(color = WGS),
    size = 3, alpha = 0.9
  ) +
  geom_point(
    data = subset(ind_coords, WGS == "yes"),
    aes(color = WGS),
    size = 3, alpha = 0.9
  ) +
  scale_color_manual(
    values = c("no" = "grey70", "yes" = "#D7191C")
  ) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "WGS"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )
```
```{r}
# Custom hex colors for the first 12 reefs
my_colors <- c(
  "#d85874", "#ce4f35", "#c78048", "#6db543", "#3e8048",
  "#63be84", "#43bdc4", "#6486c9", "#7862cc", "#cb87c9",
  "#c34cae", "#9e496b"
)

# Reef order (14 reefs total)
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
                "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
                "Wreck", "Cato", "Elizabeth", "Middleton")

# Assign the custom colors to the first 12 reefs
custom_palette <- setNames(my_colors, reef_order[1:12])

# Assign custom colors to Elizabeth and Middleton
custom_palette["Elizabeth"] <- "#c4af46"
custom_palette["Middleton"] <- "#7a7932"

# Final named palette
reef_palette <- custom_palette

# View or use reef_palette
reef_palette
```

##1.9 Plot PCA by Reef
```{r}

ggplot(ind_coords, aes(x = Axis1, y = Axis2, color = Reef)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = reef_palette) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

save(ind_coords, file = "ind_coords.RData")
```

```{r}
ind_coords_acro <- ind_coords %>%
  mutate(HostCluster = case_when(
    Axis1 < -60 ~ "AC1",
    Axis1 > -60 & Axis1 < -10 ~ "AC1b",
    Axis1 > -10 ~ "AC2",
    # Axis1 > -10 & Axis2 > 0 ~ "AC2",
    # Axis1 > 0 & Axis2 < 0 & Axis2 > -10 ~ "AC2b",
    # Axis1 > 0 & Axis2 < -10 ~ "AC3",
    TRUE ~ NA_character_  # anything that doesn't match the above
  ))

# ind_coords_acro <- ind_coords %>%
#   mutate(HostCluster = if_else(Axis1 < -80 & Axis2 > -3, "LHCluster", "CSCluster"))


save(ind_coords_acro, file = "ind_coords_acro.RData")
```


```{r}
ind_coords_acro %>%
  group_by(HostCluster) %>%
  summarise(n = n())
```

```{r}
pca.host.pop <- ggplot(ind_coords_acro, aes(x = Axis1, y = Axis2, fill = Reef)) +
  geom_point(size = 3, alpha = 0.9, color = "black") +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  scale_fill_manual(values = reef_palette) +
  theme_minimal()

pca.host.pop
```

Creating a cluster based PCA
```{r}
library(wesanderson)
library(ggplot2)
library(wesanderson)
# palette <- wes_palette("GrandBudapest1", n = 6, type = "discrete")
pca.host.pop.clusters <- ggplot(ind_coords_acro, aes(x = Axis1, y = Axis2, colour = HostCluster)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  # scale_fill_manual(values = palette) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

pca.host.pop.clusters
```

```{r}
ggsave(pca.host.pop.clusters, file = "acropora.pca.clusters.PDF",
       unit = "mm",
       height = 189,
       width = 189)
ggsave(pca.host.pop.clusters, file = "acropora.pca.clusters.PNG",
       unit = "mm",
       height = 189,
       width = 189)
```


```{r}
ind_coords_acro <- ind_coords_acro %>% filter(Reef != "NA")
ind_coords_acro <- ind_coords_acro %>% filter(Reef != "GBR")



pca.host.pop <- ggplot(ind_coords_acro, aes(x = Axis1, y = Axis2, fill = Reef, shape = HostCluster)) +
  geom_point(size = 3, alpha = 0.9, color = "black") +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  scale_fill_manual(values = reef_palette) +
    scale_shape_manual(values = c(21, 22, 23, 24, 25)) +
  guides(
    fill = guide_legend(override.aes = list(shape = 21, color = "black")),
    shape = guide_legend(override.aes = list(fill = "grey70"))  # optional shape legend tweak
  ) +
  theme_minimal()

pca.host.pop
```

```{r}
library(plotly)

# add tooltip aesthetics
pca.symbiont <- ggplot(ind_coords_acro, 
                       aes(x = Axis1, y = Axis2, 
                           fill = Reef, 
                           shape = HostCluster,
                           text = paste0(
                             "Sample: ", SampleID, "<br>",
                             "Population: ", Reef, "<br>",
                             "Host Cluster: ", HostCluster
                           ))) +
  geom_point(size = 3, alpha = 0.9, color = "black") +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  scale_fill_manual(values = reef_palette) +
  scale_shape_manual(values = c(21, 22, 23, 24, 25)) +
  guides(
    fill = guide_legend(override.aes = list(shape = 21, color = "black")),
    shape = guide_legend(override.aes = list(fill = "grey70"))
  ) +
  theme_minimal()

# make interactive with custom tooltip
pca.symbiont.interactive <- ggplotly(pca.symbiont, tooltip = "text")

pca.symbiont.interactive
```



#2. Separating datasets

##2.1 Create dataset for LH and CS
```{r}
load("acro_gen_full.RData")
load("ind_coords.RData")

# ind_coords_acro <- ind_coords %>%
#   mutate(AcroID = if_else(Axis1 < -80 & Axis2 > -3, "LHCluster", "CSCluster"))
ind_coords_acro <- ind_coords %>%
  mutate(HostCluster = case_when(
    Axis1 < -60 ~ "AC1",
    Axis1 > -60 & Axis1 < -10 ~ "AC1b",
    Axis1 > -10 ~ "AC2",
    # Axis1 > -10 & Axis2 > 0 ~ "AC2",
    # Axis1 > 0 & Axis2 < 0 & Axis2 > -10 ~ "AC2b",
    # Axis1 > 0 & Axis2 < -10 ~ "AC3",
    TRUE ~ NA_character_  # anything that doesn't match the above
  ))

# Create vectors of SampleIDs for each species based on AcroID
inds_AC1 <- ind_coords_acro$SampleID[ind_coords_acro$HostCluster == "AC1"]
inds_AC2 <- ind_coords_acro$SampleID[ind_coords_acro$HostCluster == "AC2"]

# Subset the genind object for each species
acro_gen_cl1 <- acro_gen_full[inds_AC1, ]
acro_gen_cl2 <- acro_gen_full[inds_AC2, ]

save(acro_gen_cl1, file = "acro_gen_cl1.RData")
save(acro_gen_cl2, file = "acro_gen_cl2.RData")

```

##2.2 Run PCA for LH and CS 
```{r}
# PCA for Lord Howe
x_cl1 <- tab(acro_gen_cl1 , NA.method = "mean") #issue that PCA requires non-zero variance, and suggests that loci are monomorphic.

#for PCA to work the var cannot be 0
var_cols <- apply(x_cl1, 2, var, na.rm = TRUE)
sum(var_cols == 0, na.rm=TRUE)  # monomorphic loci
# would be removing 10,234 loci 

# remove loci with zero variance (to get PCA to work)
x_cl1 <- x_cl1[, !is.na(var_cols) & var_cols > 0]

pca_cl1 <- dudi.pca(x_cl1, scannf = FALSE, scale = FALSE, nf = 3)
# # check for remaining NA 
# sum(is.na(x_cl1))

# PCA for Coral Sea 
x_cl2 <- tab(acro_gen_cl2, NA.method = "mean")
pca_cl2 <- dudi.pca(x_cl2, scannf = FALSE, scale = FALSE, nf = 3)

```

##2.3 Merge coords and meta
```{r}
#load meta ordered
meta_ordered <- read.csv("meta_ordered.csv")
# For Lord Howe 
ind_coords_cl1 <- as.data.frame(pca_cl1$li)
ind_coords_cl1$SampleID <- rownames(ind_coords_cl1)
ind_coords_cl1 <- left_join(ind_coords_cl1, meta_ordered, by = "SampleID")
save(ind_coords_cl1, file = "ind_coords_cl1.RData")
# For Coral Sea 
ind_coords_cl2 <- as.data.frame(pca_cl2$li)
ind_coords_cl2$SampleID <- rownames(ind_coords_cl2)
ind_coords_cl2 <- left_join(ind_coords_cl2, meta_ordered, by = "SampleID")
save(ind_coords_cl2, file = "ind_coords_cl2.RData")
```


```{r}
# Custom hex colors for the first 12 reefs
my_colors <- c(
  "#d85874", "#ce4f35", "#c78048", "#6db543", "#3e8048",
  "#63be84", "#43bdc4", "#6486c9", "#7862cc", "#cb87c9",
  "#c34cae", "#9e496b"
)

# Reef order (14 reefs total)
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
                "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
                "Wreck", "Cato", "Elizabeth", "Middleton")

# Assign the custom colors to the first 12 reefs
custom_palette <- setNames(my_colors, reef_order[1:12])

# Assign custom colors to Elizabeth and Middleton
custom_palette["Elizabeth"] <- "#c4af46"
custom_palette["Middleton"] <- "#7a7932"

# Final named palette
reef_palette <- custom_palette

# View or use reef_palette
reef_palette
```

##2.4 Lord Howe PCA - reef
```{r}
library(tidyverse)
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
  "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
  "Wreck", "Cato", "GBR", "Elizabeth", "Middleton")

ind_coords_cl1$Reef <- factor(ind_coords_cl1$Reef, levels = reef_order)

ind_coords_cl1 <- ind_coords_cl1 %>% filter(Reef != "NA") %>% filter(Reef != "GBR") %>% filter(Reef != "Frederick") %>% filter(Reef != "Wreck")

pca_cl1 <- ggplot(ind_coords_cl1, aes(x = Axis1, y = Axis2, color = Reef)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca_cl1$eig[1] / sum(pca_cl1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca_cl1$eig[2] / sum(pca_cl1$eig) * 100, 1), "%)"),
    color = "Reef"
  ) +
  scale_color_manual(values = reef_palette) +   
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

pca_cl1
```


##2.6 Coral Sea PCA - reef
```{r}
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
  "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
  "Wreck", "Cato", "GBR", "Elizabeth", "Middleton")

ind_coords_cl2$Reef <- factor(ind_coords_cl2$Reef, levels = reef_order)

ind_coords_cl2 <- ind_coords_cl2 %>% filter(Reef != "NA")

pca_cl2 <- ggplot(ind_coords_cl2, aes(x = Axis1, y = Axis2, color = Reef)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca_cl2$eig[1] / sum(pca_cl2$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca_cl2$eig[2] / sum(pca_cl2$eig) * 100, 1), "%)"),
    color = "Reef"
  ) +
  scale_color_manual(values = reef_palette) + 
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

pca_cl2
```


```{r}
# Add missingness to the PCA dataframe
missing_per_ind <- readRDS("missing_per_ind.RData")
ind_coords_cl2$Missingness <- missing_per_ind[ind_coords_cl2$SampleID]
```


```{r}
pca_missing <- ggplot(ind_coords_cl2, aes(x = Axis1, y = Axis2, color = Missingness)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_viridis_c(option = "plasma", end = 0.9, na.value = "grey80") +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "% Missing Data"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

pca_missing

```

# Section 3. Genetic Diversity Stats

```{r}
# packages you'll need
library(tidyverse)
library(vroom)      # fast reading
library(vcfR)       # if working with VCF directly
library(adegenet)   # PCA on genotypes (if using genind)
library(lfmm)       # or lfmm2 (install from CRAN/Bioconductor)
library(vegan)      # RDA / partial RDA
library(car)        # vif
library(qqman)      # manhattan
library(hierfstat)
#MW: loaded as part of tidyverse; library(dplyr)
# MW this probably should be loaded at the top if we want all used packages to be loaded here. library(RColorBrewer)
library(ade4)
library(MASS)      # for kde2d
library(scales) 
```

```{r}
# Load genind data
```{r}
load("acro_gen_full.RData")
load("acro_gen_cl1.RData")
load("acro_gen_cl2.RData")
load("ind_coords_acro.RData")
load("ind_coords_cl2.RData")
load("ind_coords_cl1.RData")

acro_gen_cl1
acro_gen_cl2
```

### 1. Full data
```{r}
# Let's say you have a vector of populations for each individual
meta_trimmed <- read.csv("meta_trimmed.csv")
meta_trimmed <- meta_trimmed %>% 
  mutate(MarinePark = case_when(
    Reef %in% c("Elizabeth", "Middleton") ~ "LH",
    !Reef %in% c("Elizabeth", "Middleton") ~ "CS"
  ))

ind_coords_acro <- ind_coords_acro %>% 
  mutate(MarinePark = case_when(
    Reef %in% c("Elizabeth", "Middleton") ~ "LH",
    !Reef %in% c("Elizabeth", "Middleton") ~ "CS"
  ))

#AC1
ind_coords_cl1 <- ind_coords_cl1 %>% 
  mutate(MarinePark = case_when(
    Reef %in% c("Elizabeth", "Middleton") ~ "LH",
    !Reef %in% c("Elizabeth", "Middleton") ~ "CS"
  ))
acro_gen_cl1 <- acro_gen_cl1[!indNames(acro_gen_cl1) %in% "A249-24b"] #remove extra individual from GenID
pop_vector_cl1 <- ind_coords_cl1$MarinePark # or however you're storing population info
pop(acro_gen_cl1) <- as.factor(pop_vector_cl1)

#AC2
ind_coords_cl2 <- ind_coords_cl2 %>% 
  mutate(MarinePark = case_when(
    Reef %in% c("Elizabeth", "Middleton") ~ "LH",
    !Reef %in% c("Elizabeth", "Middleton") ~ "CS"
  ))
acro_gen_cl2 <- acro_gen_cl2[
  !indNames(acro_gen_cl2) %in% c("A131-24b", "A294-24b")
] #remove the samples with b
pop_vector_cl2 <- ind_coords_cl2$MarinePark # or however you're storing population info
pop(acro_gen_cl2) <- as.factor(pop_vector_cl2)
```

## 1.1 Heterozygosity (Ho)
```{r}
pop_vector <- ind_coords_acro$HostCluster # or however you're storing population info
pop(acro_gen_full) <- as.factor(pop_vector)
# acro_gen_full <- acro_gen_full[pop(acro_gen_full) != "GBR", ]
# pop_vector <- ind_coords_acro$HostCluster # or however you're storing population info
# pop(acro_gen_full) <- as.factor(pop_vector)

#Here we are calculating basic stats using hierfstat
basic_acro <- basic.stats(acro_gen_full, diploid = TRUE)

# Mean observed heterozygosity per site
Ho_acro = apply(basic_acro$Ho, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
Ho_acro
```

## 1.2 Heterozygosity (He)
```{r}
# Mean expected heterozygosity per site
He_acro = apply(basic_acro$Hs, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
He_acro
```

#Mean Ho and He across populations
```{r}
# Mean Ho and He per population
mean_Ho <- apply(basic_acro$Ho, 2, mean, na.rm = TRUE)
mean_He <- apply(basic_acro$Hs, 2, mean, na.rm = TRUE)

round(cbind(mean_Ho, mean_He), 3)
```


##1.3 Inbreeding 
```{r}
apply(basic_acro$Fis, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
```

# AC1
## 1.1 Heterozygosity (Ho)
```{r}
#Here we are calculating basic stats using hierfstat
basic_acro <- basic.stats(acro_gen_cl1, diploid = TRUE)

# Mean observed heterozygosity per site
Ho_acro = apply(basic_acro$Ho, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
Ho_acro
```

## 1.2 Heterozygosity (He)
```{r}
# Mean expected heterozygosity per site
He_acro = apply(basic_acro$Hs, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
He_acro
```

#Mean Ho and He across populations
```{r}
# Mean Ho and He per population
mean_Ho <- apply(basic_acro$Ho, 2, mean, na.rm = TRUE)
mean_He <- apply(basic_acro$Hs, 2, mean, na.rm = TRUE)

round(cbind(mean_Ho, mean_He), 3)
```


##1.3 Inbreeding 
```{r}
apply(basic_acro$Fis, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
```

```{r}
library(ggplot2)
library(tidyr)

fis_df <- as.data.frame(basic_acro$Fis)
fis_long <- pivot_longer(fis_df, cols = everything(), names_to = "Population", values_to = "Fis")

ggplot(fis_long, aes(x = Fis)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  facet_wrap(~ Population, scales = "free_y") +
  theme_minimal() +
  labs(title = "Distribution of per-locus FIS per population")
```


##2.6 Calculate Fst
```{r}
#using hierfstat package here. 
#compute pairwise Fst values. Returns pairwise distances following Weir & Cockerham 1984 
acro_fst_cl1 <-  genet.dist(acro_gen_cl1, method = "WC84") %>% round(digits = 3)
acro_fst_cl1

save(acro_fst_cl1, file = "acro_fst_cl2.RData")
```


#AC2
## 1.1 Heterozygosity (Ho)
```{r}
#Here we are calculating basic stats using hierfstat
basic_acro <- basic.stats(acro_gen_cl2, diploid = TRUE)

# Mean observed heterozygosity per site
Ho_acro = apply(basic_acro$Ho, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
Ho_acro
```

## 1.2 Heterozygosity (He)
```{r}
# Mean expected heterozygosity per site
He_acro = apply(basic_acro$Hs, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
He_acro
```

#Mean Ho and He across populations
```{r}
# Mean Ho and He per population
mean_Ho <- apply(basic_acro$Ho, 2, mean, na.rm = TRUE)
mean_He <- apply(basic_acro$Hs, 2, mean, na.rm = TRUE)

round(cbind(mean_Ho, mean_He), 3)
```


##1.3 Inbreeding 
```{r}
apply(basic_acro$Fis, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
```
##2.6 Calculate Fst
```{r}
#using hierfstat package here. 
#compute pairwise Fst values. Returns pairwise distances following Weir & Cockerham 1984 
acro_fst_cl2 <-  genet.dist(acro_gen_cl2, method = "WC84") %>% round(digits = 3)
acro_fst_cl2

save(acro_fst_cl2, file = "acro_fst_cl2.RData")
```


```{r}
library(ggplot2)
library(tidyr)

fis_df <- as.data.frame(basic_acro$Fis)
fis_long <- pivot_longer(fis_df, cols = everything(), names_to = "Population", values_to = "Fis")

ggplot(fis_long, aes(x = Fis)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  facet_wrap(~ Population, scales = "free_y") +
  theme_minimal() +
  labs(title = "Distribution of per-locus FIS per population")
```

