---
title: "02_Isopora"
author: "Magena Marzonie and Monique White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Data Wrangle

## Load libraries
```{r}
library(dplyr)
library(vcfR)
library(adegenet)
library(data.table)
```


## Step 1: Load VCF and fix sample names 
## Load data
```{r}
vcf <- read.vcfR("ISOPORA_variants.vcf")

# Clean sample names
sample_names <- colnames(vcf@gt)[-1]
shortened_names <- sub("^([^-]+-[^-]+).*", "\\1", sample_names)
colnames(vcf@gt)[-1] <- shortened_names
```


##1. QC Filtering - all samples
###1.1 Convert genotypes to numeric format
```{r}
gt <- extract.gt(vcf, element = "GT", as.numeric = FALSE)
geno_num <- matrix(NA, nrow = nrow(gt), ncol = ncol(gt))

for (i in 1:nrow(gt)) {
  for (j in 1:ncol(gt)) {
    g <- gt[i, j]
    geno_num[i, j] <- ifelse(g %in% c("0/0", "0|0"), 0,
                       ifelse(g %in% c("0/1", "1/0", "0|1", "1|0"), 1,
                       ifelse(g %in% c("1/1", "1|1"), 2, NA)))
  }
}
```


###1.2 Loci filtering
```{r}
# Locus missingness
missing_per_locus <- rowMeans(is.na(geno_num))

# Minor allele count (MAC)
mac <- apply(geno_num, 1, function(x) {
  x <- x[!is.na(x)]
  # count whichever allele is rarer
  min(sum(x), 2*length(x) - sum(x))
})

# Keep loci: missingness ≤ 20%, MAC ≥ 3 (adjust threshold as needed)
keep_loci <- which(missing_per_locus <= 0.2 & mac >= 3)

# Apply locus filter
vcf_filtered <- vcf[keep_loci, ]

# Individual filtering
gt_filtered <- extract.gt(vcf_filtered, element = "GT", as.numeric = FALSE)
missing_per_ind <- colMeans(is.na(gt_filtered))
keep_inds <- names(missing_per_ind[missing_per_ind <= 0.5])
vcf_filtered <- vcf_filtered[, c(TRUE, colnames(gt_filtered) %in% keep_inds)]

# Remove 2020 individuals
keep_inds <- colnames(vcf_filtered)[-1]
vcf_filtered <- vcf_filtered[, c(TRUE, !grepl("-20", keep_inds))]
```


###1.3 Genind conversion / Joining metadata
```{r}
# Convert VCF to genind
iso_gen <- vcfR2genind(vcf_filtered)

# Keep only up to "-24" in individual names
indNames(iso_gen) <- sub("(-24).*", "\\1", indNames(iso_gen))

# Remove individuals with '-20' or '-24b'
inds_filtered <- indNames(iso_gen)[!grepl("-20|-24b", indNames(iso_gen))]
iso_gen <- iso_gen[inds_filtered, ]

# Read metadata
meta <- fread("Iso_Meta2024.csv")

# Clean metadata sample IDs (keep only up to "-24")
meta_clean <- meta %>%
  mutate(SampleID = sub("(-24).*", "\\1", trimws(as.character(VialID))))

# Filter and order metadata to match cleaned genind individuals
meta_ordered <- meta_clean %>%
  filter(SampleID %in% indNames(iso_gen)) %>%
  arrange(match(SampleID, indNames(iso_gen)))
```


```{r}
iso_gen <- vcfR2genind(vcf_filtered)

# Check for match
inds <- indNames(iso_gen)
inds_clean <- trimws(inds)
meta_clean <- meta %>% mutate(SampleID = trimws(as.character(VialID)))

write.csv(x = meta_clean, file ="meta_clean.csv")

# Filter out '-20' samples from inds
inds_filtered <- inds_clean[!grepl("-20|-24b", inds_clean)]

# Filter metadata to match and remove '-20' individuals
meta_ordered_iso <- meta_clean %>%
  filter(SampleID %in% inds_filtered) %>%
  arrange(match(SampleID, inds_filtered))

# Also update genind object to remove those individuals
iso_gen_full <- iso_gen[!grepl("-20|-24b", indNames(iso_gen)), ]

# #meta trimmed
# iso_meta <- read.csv("Acro_Meta2024.csv") %>%
#   rename(SampleID = VialID)
# 
# # Match and merge metadata
# inds <- indNames(acro_gen_full)
# meta_trimmed <- merge(data.table(SampleID = inds), acro_meta, by = "SampleID", all.x = TRUE)
# meta_trimmed <- meta_trimmed[match(inds, meta_trimmed$SampleID), ]

```


###Add population level assignments 
```{r}
# 3. Ensure metadata rows are ordered to match genind individuals
# indNames() gives the individual IDs in the genind
meta <- meta_ordered_iso %>%
  filter(SampleID %in% indNames(iso_gen_full)) %>%   # keep only samples in genind
  arrange(match(SampleID, indNames(iso_gen_full)))   # reorder to match genind

# 4. Assign population/reef to genind
pop_vector <- meta$Reef  
pop(iso_gen_full) <- as.factor(pop_vector)

# 5. Verify
table(pop(iso_gen_full))  # should show number of individuals per reef

meta_trimmed <- meta
save(meta_trimmed, file = "meta_trimmed.RData")
```


```{r}
save(iso_gen_full, file = "iso_gen_full.RData")

load("iso_gen_full.RData")
```

```{r}
coords <- meta[match(indNames(iso_gen_full), meta$SampleID), c("GPS S", "GPS E")]
```

```{r}
write.table(coords,
            file = "coords.txt",
            sep = "\t",
            quote = FALSE,
            row.names = TRUE,
            col.names = FALSE)
```



```{r}
# Read raw lines
lines <- readLines("geno_matrix.txt")
length(lines)      # number of rows

# Check the number of columns per line
col_counts <- sapply(lines, function(x) length(strsplit(x, "\\s+")[[1]]))
table(col_counts)  # see the distribution of column counts
```

```{r}
save(iso_gen, file = "iso_gen.RData")
save(meta_ordered_iso, file = "meta_ordered_iso.RData")
```

# Section 2. PCA

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(vcfR)
library(adegenet)
library(data.table)
```

#Step 8: Save filtered data
```{r}
# Save the filtered genind object
#save(iso_gen, file = "iso_gen.RData")

# View the genind object
load("iso_gen_full.RData")
iso_gen_full

#189 individuals, 6,583 loci, 13,293 alleles 
#187 individuals, 12,380 loci, 25,001 alleles(MW 20/4/26)
```

##1.4 PCA of genetic variance
```{r}
x <- tab(iso_gen_full, NA.method = "mean")
pca1 <- dudi.pca(x, scannf = FALSE, scale = FALSE, nf = 3)

percent <- pca1$eig / sum(pca1$eig) * 100
barplot(percent, ylab = "Variance explained (%)", names.arg = round(percent, 1))
```

##1.5 Prepare PCA coordinates + add population info
```{r}
meta_clean <- read.csv("meta_Clean.csv")
# Extract PCA coordinates
ind_coords <- as.data.frame(pca1$li)

# Add individual names
ind_coords$ind <- rownames(ind_coords)

# Add population info directly from genind object
ind_coords$pop <- iso_gen_full@pop  # already filtered to match individuals

ind_coords <- ind_coords %>% rename(SampleID = ind)

ind_coords$SampleID <- gsub("\\.", "-", ind_coords$SampleID)

ind_coords_iso <- ind_coords %>% left_join(meta_clean, by = "SampleID")


save(ind_coords_iso, file = "ind_coords_iso.RData")
```


##1.7 Plot PCA by latitude
```{r}

load("ind_coords_iso.RData")

library(wesanderson)
library(ggplot2)
library(wesanderson)

# Pick your fave Wes Anderson palette (e.g., "Darjeeling1", "Zissou1", "Moonrise1", etc.)
palette <- wes_palette("Zissou1", n = 100, type = "continuous")

ggplot(ind_coords_iso, aes(x = Axis1, y = Axis2, color = `GPS.S`)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_gradientn(colors = palette) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "Latitude"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

```


```{r}
wgs_samples <- c(
"I015-24", "I034-24", "I066-24", "I205-24", "I209-24", "I113-24", "I095-24", "I171-24", "I214-24", "I053-24", "I108-24", "I126-24"
)

library(dplyr)

ind_coords <- ind_coords %>%
  mutate(WGS_group = if_else(SampleID %in% wgs_samples,
                             SampleID,
                             "Other"))

ind_coords <- ind_coords %>%
  mutate(
    WGS_flag = SampleID %in% wgs_samples,
    SampleID_ordered = factor(SampleID, levels = wgs_samples)
  )
```

```{r}
library(ggplot2)
library(ggrepel)

ggplot(ind_coords, aes(x = Axis1, y = Axis2)) +
  
  # Grey background first
  geom_point(
    data = subset(ind_coords, !WGS_flag),
    color = "grey85",
    size = 3,
    alpha = 0.6
  ) +
  
  # WGS samples on top (default ggplot palette)
  geom_point(
    data = subset(ind_coords, WGS_flag),
    aes(color = SampleID_ordered),
    size = 4
  ) +
  
  # Labels
  geom_text_repel(
    data = subset(ind_coords, WGS_flag),
    aes(label = SampleID, color = SampleID_ordered),
    size = 3.5,
    show.legend = FALSE
  ) +
  
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "WGS Sample"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )
```


```{r}
reef_order <- c(
  "Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
  "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
  "Wreck", "Cato", "GBR", "Elizabeth", "Middleton"
)

ind_coords_iso$Reef <- factor(ind_coords_iso$Reef, levels = reef_order)

ind_coords_iso <- ind_coords_iso %>% 
  filter(Reef != "NA")

ggplot(ind_coords_iso, aes(x = Axis1, y = Axis2, color = Reef)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "Reef"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

```



```{r}
#adding host cluster so we can add shape by host 
ind_coords_iso <- ind_coords_iso %>% mutate(HostCluster = case_when(
    Axis1 < 0 & Axis2 < -10 ~ "IC1",
    Axis1 < 0 & Axis2 < 0 & Axis2 > -10 ~ "IC1b",
    Axis1 > 0 & Axis2 > 25 ~ "IC2",
    Axis1 > 20 & Axis2 < 0 ~ "IC3",
    TRUE ~ NA_character_  # anything that doesn't match the above
  ))
```


##2.1 Create dataset for LH and CS
```{r}
library(dplyr)

load("ind_coords_iso.RData")

# ind_coords_iso <- ind_coords %>%
#   mutate(isoID = if_else(Axis1 < -80 & Axis2 > -3, "LHCluster", "CSCluster"))
ind_coords_iso <- ind_coords_iso %>% mutate(HostCluster = case_when(
    Axis1 < 0 & Axis2 < 0 ~ "IC1",
    Axis1 < 0 & Axis2 < 0 ~ "IC1",
    Axis1 > 0 & Axis2 > 25 ~ "IC2",
    Axis1 > 20 & Axis2 < 0 ~ "IC3",
    TRUE ~ NA_character_  # anything that doesn't match the above
  ))

# Create vectors of SampleIDs for each species based on isoID
inds_IC1 <- ind_coords_iso$SampleID[ind_coords_iso$HostCluster == "IC1"]
inds_IC2 <- ind_coords_iso$SampleID[ind_coords_iso$HostCluster == "IC2"]
inds_IC3 <- ind_coords_iso$SampleID[ind_coords_iso$HostCluster == "IC3"]

# Subset the genind object for each species
iso_gen_cl1 <- iso_gen_full[inds_IC1, ]
iso_gen_cl2 <- iso_gen_full[inds_IC2, ]
iso_gen_cl3 <- iso_gen_full[inds_IC3, ]


save(iso_gen_cl1, file = "iso_gen_cl1.RData")
save(iso_gen_cl2, file = "iso_gen_cl2.RData")
save(iso_gen_cl3, file = "iso_gen_cl3.RData")

```
```{r}
# Custom hex colors for the first 12 reefs
my_colors <- c(
  "#d85874", "#ce4f35", "#c78048", "#6db543", "#3e8048",
  "#63be84", "#43bdc4", "#6486c9", "#7862cc", "#cb87c9",
  "#c34cae", "#9e496b"
)

# Reef order (14 reefs total)
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
                "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
                "Wreck", "Cato", "Elizabeth", "Middleton")

# Assign the custom colors to the first 12 reefs
custom_palette <- setNames(my_colors, reef_order[1:12])

# Assign custom colors to Elizabeth and Middleton
custom_palette["Elizabeth"] <- "#c4af46"
custom_palette["Middleton"] <- "#7a7932"

# Final named palette
reef_palette <- custom_palette

# View or use reef_palette
reef_palette
```

```{r}
ind_coords_iso$Reef <- factor(ind_coords_iso$Reef, levels = reef_order)
ind_coords_iso <- ind_coords_iso %>% filter(Reef != "NA")
ind_coords_iso <- ind_coords_iso %>% filter(Reef != "GBR")

iso_cluster <-  ggplot(ind_coords_iso, aes(x = Axis1, y = Axis2, colour = HostCluster)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  theme_minimal()
iso_cluster
```

```{r}
 ggsave(iso_cluster,
        file = "iso_PCA_cluster.PNG",
        height = 200,
        width = 189,
        unit = "mm")
```


#PCA for symbionts
```{r}
ind_coords_iso$Reef <- factor(ind_coords_iso$Reef, levels = reef_order)
ind_coords_iso <- ind_coords_iso %>% filter(Reef != "NA")
ind_coords_iso <- ind_coords_iso %>% filter(HostCluster != "NA")



pca.symbiont <- ggplot(ind_coords_iso, aes(x = Axis1, y = Axis2, fill = Reef, shape = HostCluster)) +
  geom_point(size = 3, alpha = 0.9, color = "black") +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  scale_fill_manual(values = reef_palette) +
    scale_shape_manual(values = c(21, 22, 23, 24, 25)) +
  guides(
    fill = guide_legend(override.aes = list(shape = 21, color = "black")),
    shape = guide_legend(override.aes = list(fill = "grey70"))  # optional shape legend tweak
  ) +
  theme_minimal()
  

pca.symbiont

```

##2.2 Run PCA for LH and CS 
```{r}
# PCA for Lord Howe
x_cl1 <- tab(iso_gen_cl1 , NA.method = "mean") #issue that PCA requires non-zero variance, and suggests that loci are monomorphic.

#for PCA to work the var cannot be 0
var_cols <- apply(x_cl1, 2, var, na.rm = TRUE)
sum(var_cols == 0, na.rm=TRUE)  # monomorphic loci


# remove loci with zero variance (to get PCA to work)
x_cl1 <- x_cl1[, !is.na(var_cols) & var_cols > 0]

pca_cl1 <- dudi.pca(x_cl1, scannf = FALSE, scale = FALSE, nf = 3)
# # check for remaining NA 
# sum(is.na(x_cl1))

# PCA for Coral Sea 
x_cl2 <- tab(iso_gen_cl2, NA.method = "mean")
pca_cl2 <- dudi.pca(x_cl2, scannf = FALSE, scale = FALSE, nf = 3)


# IC3
x_cl3 <- tab(iso_gen_cl3, NA.method = "mean")
pca_cl3 <- dudi.pca(x_cl3, scannf = FALSE, scale = FALSE, nf = 3)
```
##2.3 Merge coords and meta
```{r}
#load meta ordered
meta_ordered_iso <- load("meta_ordered_iso.RData")

# For CL1
ind_coords_cl1 <- as.data.frame(pca_cl1$li)
ind_coords_cl1$SampleID <- rownames(ind_coords_cl1)
ind_coords_cl1 <- left_join(ind_coords_cl1, meta_ordered, by = "SampleID")
save(ind_coords_cl1, file = "ind_coords_cl1.RData")
# For CL2
ind_coords_cl2 <- as.data.frame(pca_cl2$li)
ind_coords_cl2$SampleID <- rownames(ind_coords_cl2)
ind_coords_cl2 <- left_join(ind_coords_cl2, meta_ordered, by = "SampleID")
save(ind_coords_cl2, file = "ind_coords_cl2.RData")

# CL3
ind_coords_cl3 <- as.data.frame(pca_cl3$li)
ind_coords_cl3$SampleID <- rownames(ind_coords_cl3)
ind_coords_cl3 <- left_join(ind_coords_cl3, meta_ordered, by = "SampleID")
save(ind_coords_cl3, file = "ind_coords_cl3.RData")
```
##2.4 CL1 PCA
```{r}
library(tidyverse)
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
  "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
  "Wreck", "Cato", "GBR", "Elizabeth", "Middleton")

ind_coords_cl1$Reef <- factor(ind_coords_cl1$Reef, levels = reef_order)

ind_coords_cl1 <- ind_coords_cl1 %>% filter(Reef != "NA") %>% filter(Reef != "GBR") %>% filter(Reef != "Frederick") %>% filter(Reef != "Wreck")

pca_cl1 <- ggplot(ind_coords_cl1, aes(x = Axis1, y = Axis2, color = Reef)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca_cl1$eig[1] / sum(pca_cl1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca_cl1$eig[2] / sum(pca_cl1$eig) * 100, 1), "%)"),
    color = "Reef"
  ) +
  scale_color_manual(values = reef_palette) +   
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

pca_cl1
```


##2.6 CL2
```{r}
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
  "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
  "Wreck", "Cato", "GBR", "Elizabeth", "Middleton")

ind_coords_cl2$Reef <- factor(ind_coords_cl2$Reef, levels = reef_order)

ind_coords_cl2 <- ind_coords_cl2 %>% filter(Reef != "NA")

pca_cl2 <- ggplot(ind_coords_cl2, aes(x = Axis1, y = Axis2, color = Reef)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca_cl2$eig[1] / sum(pca_cl2$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca_cl2$eig[2] / sum(pca_cl2$eig) * 100, 1), "%)"),
    color = "Reef"
  ) +
  scale_color_manual(values = reef_palette) + 
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

pca_cl2
```

#CL3
```{r}
library(tidyverse)
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
  "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
  "Wreck", "Cato", "GBR", "Elizabeth", "Middleton")

ind_coords_cl3$Reef <- factor(ind_coords_cl3$Reef, levels = reef_order)

ind_coords_cl3 <- ind_coords_cl3 %>% filter(Reef != "NA") %>% filter(Reef != "GBR") %>% filter(Reef != "Frederick") %>% filter(Reef != "Wreck")

pca_cl3 <- ggplot(ind_coords_cl3, aes(x = Axis1, y = Axis2, color = Reef)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca_cl3$eig[1] / sum(pca_cl3$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca_cl3$eig[2] / sum(pca_cl3$eig) * 100, 1), "%)"),
    color = "Reef"
  ) +
  scale_color_manual(values = reef_palette) +   
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

pca_cl3
```


```{r}
#How many total individuals 
#How many per reef
# Total count
total_individuals <- ind_coords_iso %>%
  summarise(n_total = n())

print(total_individuals)

# Count per reef
individuals_per_reef <- ind_coords_iso %>%
  group_by(Reef) %>%
  summarise(n_individuals = n()) 

print(individuals_per_reef)
```



##More investigations on missingness



Step 3: Filter loci by missing data and MAF 
```{r}
# Keep loci with ≤20% missing data
missing_per_locus <- apply(geno_num, 1, function(x) mean(is.na(x)))
keep_loci_miss <- which(missing_per_locus <= 0.2)

# Keep loci with MAF ≥ 0.05
maf <- apply(geno_num, 1, function(x) {
  x <- x[!is.na(x)]
  p <- mean(x) / 2
  min(p, 1 - p)
})
keep_loci_maf <- which(maf >= 0.05)

# Apply locus filters
keep_loci <- intersect(keep_loci_miss, keep_loci_maf)
vcf_loci_filtered <- vcf[keep_loci, ]
```


Step 4: Filter individuals by missing data
```{r}
# Re-extract the genotype matrix after filtering
gt_filtered <- extract.gt(vcf_loci_filtered, element = "GT", as.numeric = FALSE)

# Filter by missing data per individual
missing_per_ind <- apply(gt_filtered, 2, function(x) mean(is.na(x)))
keep_inds <- names(missing_per_ind[missing_per_ind <= 0.5])

# Keep only the selected individuals and FORMAT column
vcf_final <- vcf_loci_filtered[, c(TRUE, colnames(gt_filtered) %in% keep_inds)]
```


```{r}
library(tibble)
library(ggplot2)

# Create a tibble of individuals and missingness
missing_df <- tibble(
  Individual = names(missing_per_ind),
  MissingPercent = missing_per_ind * 100
)
```


```{r}
ggplot(missing_df, aes(x = MissingPercent)) +
  geom_histogram(bins = 30, fill = "tomato", color = "black") +
  theme_minimal() +
  labs(
    title = "Missing Genotype Data per Individual",
    x = "% Missing Genotypes",
    y = "Number of Individuals"
  ) +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red") +
  annotate("text", x = 52, y = max(table(cut(missing_df$MissingPercent, breaks=30))), 
           label = "50% threshold", hjust = 0, color = "red", size = 3.5)
```


```{r}
# Missing data per locus
missing_per_locus <- apply(gt_filtered, 1, function(x) mean(is.na(x)))

locus_df <- tibble(
  Locus = rownames(gt_filtered),
  MissingPercent = missing_per_locus * 100
)

ggplot(locus_df, aes(x = MissingPercent)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Missing Genotype Data per Locus",
    x = "% Missing Genotypes",
    y = "Number of Loci"
  ) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "red") +
  annotate("text", x = 12, y = max(table(cut(locus_df$MissingPercent, breaks=30))), 
           label = "10% threshold", hjust = 0, color = "red", size = 3.5)
```




```{r}
# If you're using a genind object:
missing_per_ind <- apply(is.na(iso_gen@tab), 1, mean)

# Or if you're using a genotype matrix (like from extract.gt)
missing_per_ind <- apply(gt_filtered, 2, function(x) mean(is.na(x)))

# Convert to % missing
missing_per_ind <- missing_per_ind * 100
```

```{r}
# Check that rownames match
head(rownames(ind_coords))
head(names(missing_per_ind))

# Add missingness to the PCA dataframe
ind_coords$Missingness <- missing_per_ind[ind_coords$SampleID]
```

```{r}
pca_missing <- ggplot(ind_coords, aes(x = Axis1, y = Axis2, color = Missingness)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_viridis_c(option = "plasma", end = 0.9, na.value = "grey80") +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "% Missing Data"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

pca_missing

```



```{r}
library(plotly)

# add tooltip aesthetics
pca.symbiont <- ggplot(ind_coords_iso, 
                       aes(x = Axis1, y = Axis2, 
                           fill = Reef, 
                           shape = HostCluster,
                           text = paste0(
                             "Sample: ", SampleID, "<br>",
                             "Population: ", Reef, "<br>",
                             "Host Cluster: ", HostCluster
                           ))) +
  geom_point(size = 3, alpha = 0.9, color = "black") +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  scale_fill_manual(values = reef_palette) +
  scale_shape_manual(values = c(21, 22, 23, 24, 25)) +
  guides(
    fill = guide_legend(override.aes = list(shape = 21, color = "black")),
    shape = guide_legend(override.aes = list(fill = "grey70"))
  ) +
  theme_minimal()

# make interactive with custom tooltip
pca.symbiont.interactive <- ggplotly(pca.symbiont, tooltip = "text")

pca.symbiont.interactive
```

# Section 3. Genetic Diversity Stats


# Section 4. Gene Environment Analysis
#Load libraries
```{r}
# Load libraries
library(dartR)
library(adegenet)
library(poppr)
library(hierfstat)
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(dplyr)
library(tidyverse)
library(vegan)
library(corrplot)
```


#1.1 Load genotype data
```{r}
## 1.1 Load genotype data
load("iso_gen.RData")

prep_genotype <- function(genind_obj) {

  gen <- genind_obj %>% as.matrix()

  # keep loci, not alleles
  gen <- gen[, rep(c(TRUE, FALSE), times = ncol(gen)/2)]

  # impute (modal allele)
  gen_imp <- apply(gen, 2, function(x) {
    if (any(is.na(x))) {
      replICe(x, is.na(x), as.numeric(names(which.max(table(x)))))
    } else x
  }) %>% as.data.frame()

  gen_imp
}
```

```{r}
## 1.2 Load environmental data
enviro <- read.csv("enviro_meta.csv") %>%
  rename(Reef = POI_ID) %>%
  mutate(Reef = as.fICtor(Reef)) %>%
  select(Reef, CHL_me_002, DHW_me_002, DHW_sd_002,
         SCV_me_002, PH_me_002, SSS_me_002,
         SST_me_002, SST_sd_002)

load("ind_coords_iso.RData")

ind_coords_gea <- ind_coords_iso %>% mutate(HostCluster = case_when(
    Axis1 < 0 & Axis2 < 0 ~ "IC1",
    Axis1 > 0 & Axis2 > 25 ~ "IC2",
    Axis1 > 20 & Axis2 < 0 ~ "IC3",
    TRUE ~ NA_charICter_  # anything that doesn't match the above
  ))

full.meta <- meta_clean %>%
  left_join(enviro, by = "Reef") %>%
  left_join(ind_coords_gea) %>%
  na.omit() %>%
  column_to_rownames("SampleID") %>%
  dplyr::select(-Reference, -Site, -Date, -Bag, -Species,
         -Vial, -Year, -`GPS E`, -`GPS S`) %>%
  rename(Depth = `Depth (m)`)
```

```{r}
#little sanity check here
table(full.meta$HostCluster)
#$IC1 IC2 IC3 
# 64  38  83 
```

```{r}
build_gen_meta <- function(genind_obj, meta) {

  meta$SampleID <- rownames(meta)
  ind_names     <- as.charICter(indNames(genind_obj))

  common <- intersect(ind_names, meta$SampleID)

  message("Individuals retained: ", length(common))

  genind_sub <- genind_obj[common, drop = TRUE]
  meta_sub   <- meta[common, , drop = FALSE]

  gen <- prep_genotype(genind_sub)

  message("Initial genotype matrix: ",
          nrow(gen), " inds x ", ncol(gen), " SNPs")

  ## remove invariant SNPs (CRITICAL for RDA)
  var_keep <- apply(gen, 2, var, na.rm = TRUE) > 0
  gen <- gen[, var_keep, drop = FALSE]

  message("After invariant filter: ",
          nrow(gen), " inds x ", ncol(gen), " SNPs")

  list(
    gen  = gen,
    meta = meta_sub
  )
}
```

#4. Build eICh dataset

##4.1 Full dataset
```{r}
all_data <- build_gen_meta(
  genind_obj = iso_gen,
  meta       = full.meta
)

gen_all  <- all_data$gen
meta_all <- all_data$meta

dim(gen_all)
#[1]   185 12497
```

##4.2 Host cluster IC1
```{r}
IC1_data <- build_gen_meta(
  genind_obj = iso_gen,
  meta       = full.meta %>% filter(HostCluster == "IC1")
)

gen_IC1  <- IC1_data$gen
meta_IC1 <- IC1_data$meta

dim(gen_IC1)
#   64 6936
```

##4.3 Host cluster IC2
```{r}
IC2_data <- build_gen_meta(
  genind_obj = iso_gen,
  meta       = full.meta %>% filter(HostCluster == "IC2")
)

gen_IC2  <- IC2_data$gen
meta_IC2 <- IC2_data$meta

dim(gen_IC2)
# 38 7569
```

##4.3 Host cluster IC3
```{r}
IC3_data <- build_gen_meta(
  genind_obj = iso_gen,
  meta       = full.meta %>% filter(HostCluster == "IC3")
)

gen_IC3  <- IC3_data$gen
meta_IC3 <- IC3_data$meta

dim(gen_IC3)
#[1]   83 9063
```

#5. Sanity check for all genetic matrices
```{r}
summary(apply(gen_all, 2, var))
summary(apply(gen_IC1, 2, var))
summary(apply(gen_IC2, 2, var))
summary(apply(gen_IC3, 2, var))
```

```{r}
library(vegan)

run_gea_rda <- function(gen, meta, nperm = 99) {

  ## -----------------------------
  ## 1. Fit RDA with environmental variables
  ## -----------------------------
  rda_mod <- rda(
    gen ~ Depth + DHW_me_002 + SST_sd_002 + SCV_me_002,
    data = meta
  )

  ## -----------------------------
  ## 2. Optional bICkward selection
  ## -----------------------------
  # Stepwise selection using permutations
  rda_sel <- ordistep(
    rda_mod,
    direction = "bICkward",
    permutations = nperm
  )

  ## -----------------------------
  ## 3. Test overall significance of RDA
  ## -----------------------------
  rda_overall_sig <- anova(rda_sel, permutations = nperm)

  ## -----------------------------
  ## 4. Test significance of individual enviro variables
  ## -----------------------------
  rda_terms_sig <- anova(rda_sel, by = "term", permutations = nperm)

  ## -----------------------------
  ## 5. ExtrICt SNP loadings
  ## -----------------------------
  loadings <- summary(rda_sel)$species[, 1:2, drop = FALSE]

  get_outliers <- function(x, z = 2.5) {
    lim <- mean(x) + c(-1, 1) * z * sd(x)
    names(x[x < lim[1] | x > lim[2]])
  }

  cand <- unique(c(
    get_outliers(loadings[, 1]),
    get_outliers(loadings[, 2])
  ))

  ## -----------------------------
  ## 6. Return results as a list
  ## -----------------------------
  list(
    rda           = rda_sel,        # RDA model after selection
    candidates    = cand,           # candidate SNPs
    Rsq           = RsquareAdj(rda_sel),  # adjusted R^2
    loadings      = loadings,       # SNP loadings
    anova_overall = rda_overall_sig,   # overall RDA significance
    anova_terms   = rda_terms_sig      # per-term significance
  )
}
```

```{r}
c("Depth", "DHW_me_002", "SST_sd_002", "SCV_me_002") %in% colnames(meta_all)
#make sure this returns TRUE for all 
```

```{r}
gea_all <- run_gea_rda(gen_all, meta_all)
gea_IC1 <- run_gea_rda(gen_IC1, meta_IC1)
gea_IC2 <- run_gea_rda(gen_IC2, meta_IC2)
gea_IC3 <- run_gea_rda(gen_IC3, meta_IC3)

```

```{r}
# Adjusted R²
gea_all$Rsq #0.2469014
gea_IC1$Rsq #0.1317304
gea_IC2$Rsq #0.1059666
gea_IC3$Rsq #0.09652213

# Number of candidate SNPs
length(gea_all$candidates) #1140
length(gea_IC1$candidates) #448
length(gea_IC2$candidates) #498
length(gea_IC3$candidates) # 687

# Per-variable p-values
gea_all$anova_terms
#            Df Variance       F Pr(>F)   
#Depth        1   130.07 11.1528   0.01 **
#DHW_me_002   1   443.01 37.9856   0.01 **
#SST_sd_002   1   109.18  9.3615   0.01 **
#SCV_me_002   1    67.92  5.8240   0.01 **
#Residual   180  2099.26     

gea_IC1$anova_terms
#           Df Variance      F Pr(>F)   
#SST_sd_002  1   173.62 9.1927   0.01 **
#SCV_me_002  1    44.68 2.3655   0.01 **
#Residual   61  1152.10  

gea_IC2$anova_terms
#           Df Variance      F Pr(>F)   
#Depth       1   113.06 2.3355   0.04 * 
#DHW_me_002  1    69.00 1.4254   0.12   
#SCV_me_002  1   175.46 3.6246   0.01 **
#Residual   34  1645.90         

gea_IC3$anova_terms
#           Df Variance      F Pr(>F)   
#Depth       1    40.35 2.1368   0.01 **
#DHW_me_002  1   106.32 5.6297   0.01 **
#SST_sd_002  1    55.23 2.9245   0.01 **
#SCV_me_002  1    39.08 2.0694   0.01 **
#Residual   78  1473.08       
```


```{r}
# ExtrICt candidate loci
load.rda <- summary(gea_all$rda)$species[,1:2]  # only RDA1 & RDA2
hist(load.rda[,1], main="Loadings on RDA1 (ALL)")
hist(load.rda[,2], main="Loadings on RDA2 (ALL)")

# Outliers function
outliers <- function(x, z){
  lims <- mean(x) + c(-1, 1) * z * sd(x)
  x[x < lims[1] | x > lims[2]]
}

cand1 <- outliers(load.rda[,1], 2.5)
cand2 <- outliers(load.rda[,2], 2.5)
ICro.rda.cand <- unique(c(names(cand1), names(cand2)))

meta2 <- meta_all %>% dplyr::select(Depth, DHW_me_002, SST_sd_002, SCV_me_002)

cand <- data.frame(snp = ICro.rda.cand, stringsAsFICtors = FALSE)
cand <- subset(cand, snp %in% colnames(gen_all))

cor_matrix <- matrix(nrow = nrow(cand), ncol = ncol(meta2))
colnames(cor_matrix) <- colnames(meta2)

for (i in 1:nrow(cand)) {
  snp_name <- cand$snp[i]
  snp_gen <- gen_all[, snp_name]
  cor_matrix[i, ] <- sapply(1:ncol(meta2), function(j)
    cor(as.numeric(meta2[, j]), as.numeric(snp_gen)))
}

cand <- cbind.data.frame(cand, cor_matrix)
cand <- cand[!duplicated(cand$snp),]

cand$predictor <- apply(cand[, 2:5], 1, function(x) names(which.max(abs(x))))
cand$correlation <- apply(cand[, 2:5], 1, function(x) max(abs(x)))

# Color coding
cols <- rep(adjustcolor('grey', 0.5), times = ncol(gen_all))
names(cols) <- colnames(gen_all)
cols[cand$snp[cand$predictor == 'Depth']] <- '#A7226E'
cols[cand$snp[cand$predictor == 'DHW_me_002']] <- '#EC2049'
cols[cand$snp[cand$predictor == 'SCV_me_002']] <- '#F26B38'
cols[cand$snp[cand$predictor == 'SST_sd_002']] <- '#E1F5C4'

# Plot
plot(gea_all$rda, type="n", scaling = 3, xlim = c(-1,1), ylim = c(-1,1))
points(gea_all$rda, display = "species", pch = 21, cex = 1, col = "gray32", bg = cols, scaling = 3)
text(gea_all$rda, scaling = 3, display = "bp", col = "#0868IC", cex = 1)
legend("bottomright", legend = c("Depth", "DHW mean", "SCV mean", "SST var"),  
       bty = "n", col = "gray32", pch = 21, cex = 1, pt.bg = c('#A7226E',"#EC2049", "#F26B38", '#E1F5C4'))
```


```{r}
load.rda <- summary(gea_IC1$rda)$species[,1:2]
hist(load.rda[,1], main="Loadings on RDA1 (IC1)")
hist(load.rda[,2], main="Loadings on RDA2 (IC1)")

cand1 <- outliers(load.rda[,1], 2.5)
cand2 <- outliers(load.rda[,2], 2.5)
ICro.rda.cand <- unique(c(names(cand1), names(cand2)))

meta2 <- meta_IC1 %>% dplyr::select(Depth, DHW_me_002, SST_sd_002, SCV_me_002)

cand <- data.frame(snp = ICro.rda.cand, stringsAsFICtors = FALSE)
cand <- subset(cand, snp %in% colnames(gen_IC1))

cor_matrix <- matrix(nrow = nrow(cand), ncol = ncol(meta2))
colnames(cor_matrix) <- colnames(meta2)

for (i in 1:nrow(cand)) {
  snp_name <- cand$snp[i]
  snp_gen <- gen_IC1[, snp_name]
  cor_matrix[i, ] <- sapply(1:ncol(meta2), function(j)
    cor(as.numeric(meta2[, j]), as.numeric(snp_gen)))
}

cand <- cbind.data.frame(cand, cor_matrix)
cand <- cand[!duplicated(cand$snp),]

cand$predictor <- apply(cand[, 2:5], 1, function(x) names(which.max(abs(x))))
cand$correlation <- apply(cand[, 2:5], 1, function(x) max(abs(x)))

cols <- rep(adjustcolor('grey', 0.5), times = ncol(gen_IC1))
names(cols) <- colnames(gen_IC1)
cols[cand$snp[cand$predictor == 'Depth']] <- '#A7226E'
cols[cand$snp[cand$predictor == 'DHW_me_002']] <- '#EC2049'
cols[cand$snp[cand$predictor == 'SCV_me_002']] <- '#F26B38'
cols[cand$snp[cand$predictor == 'SST_sd_002']] <- '#E1F5C4'

plot(gea_IC1$rda, type="n", scaling = 3, xlim = c(-1,1), ylim = c(-1,1))
points(gea_IC1$rda, display = "species", pch = 21, cex = 1, col = "gray32", bg = cols, scaling = 3)
text(gea_IC1$rda, scaling = 3, display = "bp", col = "#0868IC", cex = 1)
legend("bottomright", legend = c("Depth", "DHW mean", "SCV mean", "SST var"),  
       bty = "n", col = "gray32", pch = 21, cex = 1, pt.bg = c('#A7226E',"#EC2049", "#F26B38", '#E1F5C4'))
```


```{r}
load.rda <- summary(gea_IC2$rda)$species[,1:2]
hist(load.rda[,1], main="Loadings on RDA1 (IC2)")
hist(load.rda[,2], main="Loadings on RDA2 (IC2)")

cand1 <- outliers(load.rda[,1], 2.5)
cand2 <- outliers(load.rda[,2], 2.5)
ICro.rda.cand <- unique(c(names(cand1), names(cand2)))

meta2 <- meta_IC2 %>% dplyr::select(Depth, DHW_me_002, SST_sd_002, SCV_me_002)

cand <- data.frame(snp = ICro.rda.cand, stringsAsFICtors = FALSE)
cand <- subset(cand, snp %in% colnames(gen_IC2))

cor_matrix <- matrix(nrow = nrow(cand), ncol = ncol(meta2))
colnames(cor_matrix) <- colnames(meta2)

for (i in 1:nrow(cand)) {
  snp_name <- cand$snp[i]
  snp_gen <- gen_IC2[, snp_name]
  cor_matrix[i, ] <- sapply(1:ncol(meta2), function(j)
    cor(as.numeric(meta2[, j]), as.numeric(snp_gen)))
}

cand <- cbind.data.frame(cand, cor_matrix)
cand <- cand[!duplicated(cand$snp),]

cand$predictor <- apply(cand[, 2:5], 1, function(x) names(which.max(abs(x))))
cand$correlation <- apply(cand[, 2:5], 1, function(x) max(abs(x)))

cols <- rep(adjustcolor('grey', 0.5), times = ncol(gen_IC2))
names(cols) <- colnames(gen_IC2)
cols[cand$snp[cand$predictor == 'Depth']] <- '#A7226E'
cols[cand$snp[cand$predictor == 'DHW_me_002']] <- '#EC2049'
cols[cand$snp[cand$predictor == 'SCV_me_002']] <- '#F26B38'
cols[cand$snp[cand$predictor == 'SST_sd_002']] <- '#E1F5C4'

plot(gea_IC2$rda, type="n", scaling = 3, xlim = c(-1,1), ylim = c(-1,1))
points(gea_IC2$rda, display = "species", pch = 21, cex = 1, col = "gray32", bg = cols, scaling = 3)
text(gea_IC2$rda, scaling = 3, display = "bp", col = "#0868IC", cex = 1)
legend("bottomright", legend = c("Depth", "DHW mean", "SCV mean", "SST var"),  
       bty = "n", col = "gray32", pch = 21, cex = 1, pt.bg = c('#A7226E',"#EC2049", "#F26B38", '#E1F5C4'))
```

```{r}
load.rda <- summary(gea_IC3$rda)$species[,1:2]
hist(load.rda[,1], main="Loadings on RDA1 (IC3)")
hist(load.rda[,2], main="Loadings on RDA2 (IC3)")

cand1 <- outliers(load.rda[,1], 2.5)
cand2 <- outliers(load.rda[,2], 2.5)
ICro.rda.cand <- unique(c(names(cand1), names(cand2)))

meta2 <- meta_IC3 %>% dplyr::select(Depth, DHW_me_002, SST_sd_002, SCV_me_002)

cand <- data.frame(snp = ICro.rda.cand, stringsAsFICtors = FALSE)
cand <- subset(cand, snp %in% colnames(gen_IC3))

cor_matrix <- matrix(nrow = nrow(cand), ncol = ncol(meta2))
colnames(cor_matrix) <- colnames(meta2)

for (i in 1:nrow(cand)) {
  snp_name <- cand$snp[i]
  snp_gen <- gen_IC3[, snp_name]
  cor_matrix[i, ] <- sapply(1:ncol(meta2), function(j)
    cor(as.numeric(meta2[, j]), as.numeric(snp_gen)))
}

cand <- cbind.data.frame(cand, cor_matrix)
cand <- cand[!duplicated(cand$snp),]

cand$predictor <- apply(cand[, 2:5], 1, function(x) names(which.max(abs(x))))
cand$correlation <- apply(cand[, 2:5], 1, function(x) max(abs(x)))

cols <- rep(adjustcolor('grey', 0.5), times = ncol(gen_IC3))
names(cols) <- colnames(gen_IC3)
cols[cand$snp[cand$predictor == 'Depth']] <- '#A7226E'
cols[cand$snp[cand$predictor == 'DHW_me_002']] <- '#EC2049'
cols[cand$snp[cand$predictor == 'SCV_me_002']] <- '#F26B38'
cols[cand$snp[cand$predictor == 'SST_sd_002']] <- '#E1F5C4'

plot(gea_IC3$rda, type="n", scaling = 3, xlim = c(-1,1), ylim = c(-1,1))
points(gea_IC3$rda, display = "species", pch = 21, cex = 1, col = "gray32", bg = cols, scaling = 3)
text(gea_IC3$rda, scaling = 3, display = "bp", col = "#0868IC", cex = 1)
legend("bottomright", legend = c("Depth", "DHW mean", "SCV mean", "SST var"),  
       bty = "n", col = "gray32", pch = 21, cex = 1, pt.bg = c('#A7226E',"#EC2049", "#F26B38", '#E1F5C4'))
```


```{r}
library(ggplot2)
library(grid)
library(dplyr)
library(ggrepel)
library(patchwork)

# ------------------------------
# Helper function for single RDA plot
# ------------------------------
make_single_rda_plot <- function(gea_result, gen_matrix, meta_matrix, title = "", show_legend = FALSE) {

  # Candidate SNPs
  load.rda <- summary(gea_result$rda)$species[,1:2]
  outliers <- function(x, z = 2.5) {
    lims <- mean(x) + c(-1,1) * z * sd(x)
    x[x < lims[1] | x > lims[2]]
  }
  cand1 <- outliers(load.rda[,1])
  cand2 <- outliers(load.rda[,2])
  cand_snps <- unique(c(names(cand1), names(cand2)))

  meta2 <- meta_matrix %>% dplyr::select(Depth, DHW_me_002, SST_sd_002, SCV_me_002)

  cand <- data.frame(snp = cand_snps, stringsAsFICtors = FALSE)
  cand <- subset(cand, snp %in% colnames(gen_matrix))

  # Correlations
  cor_matrix <- matrix(nrow = nrow(cand), ncol = ncol(meta2))
  colnames(cor_matrix) <- colnames(meta2)
  for (i in 1:nrow(cand)) {
    snp_name <- cand$snp[i]
    snp_gen <- gen_matrix[, snp_name]
    cor_matrix[i, ] <- sapply(1:ncol(meta2), function(j) cor(as.numeric(meta2[, j]), as.numeric(snp_gen)))
  }

  cand <- cbind.data.frame(cand, cor_matrix)
  cand$predictor <- apply(cand[, 2:5], 1, function(x) names(which.max(abs(x))))

  # Color coding
  predictor_colors <- c(
    "Depth" = '#A7226E',
    "DHW_me_002" = '#EC2049',
    "SCV_me_002" = '#F26B38',
    "SST_sd_002" = '#E1F5C4'
  )
  cols <- rep(adjustcolor('grey', 0.5), times = ncol(gen_matrix))
  names(cols) <- colnames(gen_matrix)
  for (pred in names(predictor_colors)) {
    cols[cand$snp[cand$predictor == pred]] <- predictor_colors[pred]
  }

  # Scores
  species_scores <- scores(gea_result$rda, display = "species", scaling = 3)
  env_scores <- scores(gea_result$rda, display = "bp", scaling = 3)
  arrow_multiplier <- 0.8 * max(abs(species_scores)) / max(abs(env_scores))
  env_scores_scaled <- env_scores * arrow_multiplier

  species_df <- as.data.frame(species_scores) %>% mutate(Species = rownames(.), fill = cols)
  env_df <- as.data.frame(env_scores_scaled) %>% mutate(Env = rownames(.))

  # Variance explained
  var_expl <- summary(gea_result$rda)$cont$importance
  axis1_var <- round(var_expl[2,1]*100,1)
  axis2_var <- round(var_expl[2,2]*100,1)

  # Dynamic axis limits
  all_vals <- c(species_df$RDA1, species_df$RDA2, env_df$RDA1, env_df$RDA2)
  axis_lim <- max(abs(all_vals)) * 1.1  # 10% padding

  # Nudges for arrows
  env_df <- env_df %>%
    mutate(
      Env = recode(Env,
                   "DHW_me_002" = "Mean DHW",
                   "SCV_me_002" = "Mean SCV",
                   "SST_sd_002" = "Var SST",
                   "Depth" = "Depth"),
      nudge_x = ifelse(RDA1 >= 0, 0.05*axis_lim, -0.05*axis_lim),
      nudge_y = ifelse(RDA2 >= 0, 0.05*axis_lim, -0.05*axis_lim)
    )

  # Plot
  p <- ggplot() +
    geom_point(data = species_df, aes(x = RDA1, y = RDA2, fill = fill),
               shape = 21, size = 3, color = "gray32") +
    geom_segment(data = env_df, aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
                 arrow = arrow(length = unit(0.2,"cm")), color = "black", linewidth = 0.5) +
    geom_text_repel(data = env_df, aes(x = RDA1, y = RDA2, label = Env),
                    nudge_x = env_df$nudge_x, nudge_y = env_df$nudge_y,
                    color = "black", size = 4, seed = 42) +
    scale_fill_identity(
      guide = if(show_legend) "legend" else "none",
      labels = if(show_legend) c("Depth", "Mean DHW", "Mean SCV", "Var SST") else NULL,
      breaks = if(show_legend) predictor_colors else NULL,
      name = if(show_legend) "Top Predictor" else NULL
    ) +
    coord_fixed(xlim = c(-axis_lim, axis_lim), ylim = c(-axis_lim, axis_lim)) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.line = element_line(color = "black"),
          legend.position = if(show_legend) c(1.3, 1.0) else "none",
          legend.justification = c("right", "top"),
          legend.direction = "vertical",
          legend.bICkground = if(show_legend) element_rect(fill = "white", color = "black") else element_blank(),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 9)
    ) +
    xlab(paste0("RDA1 (", axis1_var, "%)")) +
    ylab(paste0("RDA2 (", axis2_var, "%)")) +
    ggtitle(title)

  return(p)
}

# ------------------------------
# Make plots
# ------------------------------
rda_all_plot <- make_single_rda_plot(gea_all, gen_all, meta_all, title = "ALL samples", show_legend = FALSE)
rda_IC1_plot  <- make_single_rda_plot(gea_IC1, gen_IC1, meta_IC1, title = "Host Cluster IC1", show_legend = FALSE)
rda_IC2_plot  <- make_single_rda_plot(gea_IC2, gen_IC2, meta_IC2, title = "Host Cluster IC2", show_legend = FALSE)
rda_IC3_plot  <- make_single_rda_plot(gea_IC3, gen_IC3, meta_IC3, title = "Host Cluster IC3", show_legend = TRUE)


# ------------------------------
# Combine side by side
# ------------------------------
library(ggplot2)
library(gridExtra)

# -------------------------------
# 1. Make legend-only plot
# -------------------------------
legend_df <- data.frame(
  Predictor = c("Depth", "Mean DHW", "Mean SCV", "Var SST"),
  x = 1:4, y = 1:4,  # dummy coordinates
  fill = c('#A7226E', '#EC2049', '#F26B38', '#E1F5C4')
)

legend_plot <- ggplot(legend_df, aes(x = x, y = y, fill = fill)) +
  geom_point(shape = 21, size = 5) +
  scale_fill_identity(name = "Top Predictor",
                      labels = legend_df$Predictor,
                      guide = "legend") +
  theme_void() +
  theme(
    legend.position = "right",
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12, face = "bold")
  )

# -------------------------------
# 2. Remove legends from RDA plots
# -------------------------------
rda_all_plot_noleg <- rda_all_plot + theme(legend.position = "none")
rda_IC1_plot_noleg <- rda_IC1_plot + theme(legend.position = "none")
rda_IC2_plot_noleg <- rda_IC2_plot + theme(legend.position = "none")
rda_IC3_plot_noleg <- rda_IC3_plot + theme(legend.position = "none")


# -------------------------------
# 3. Arrange plots side by side with legend
# -------------------------------
grid.arrange(
  rda_all_plot_noleg,
  rda_IC1_plot_noleg,
  rda_IC2_plot_noleg,
  rda_IC3_plot_noleg,
  legend_plot,
  ncol = 5,
  widths = c(1,1,1,1,0.4)  # make legend narrower
)
```



