---
title: "04_P_damicornis"
author: "Magena Marzonie and Monique White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Data Wrangle
#Load packages
```{r}
library(dplyr)
library(vcfR)
library(adegenet)
library(data.table)
```



```{r}
vcf <- read.vcfR("pdam_variants.vcf")

# Clean sample names
sample_names <- colnames(vcf@gt)[-1]
shortened_names <- sub("^([^-]+-[^-]+).*", "\\1", sample_names)
colnames(vcf@gt)[-1] <- shortened_names
```


```{r}
gt <- extract.gt(vcf, element = "GT", as.numeric = FALSE)
geno_num <- matrix(NA, nrow = nrow(gt), ncol = ncol(gt))

for (i in 1:nrow(gt)) {
  for (j in 1:ncol(gt)) {
    g <- gt[i, j]
    geno_num[i, j] <- ifelse(g %in% c("0/0", "0|0"), 0,
                       ifelse(g %in% c("0/1", "1/0", "0|1", "1|0"), 1,
                       ifelse(g %in% c("1/1", "1|1"), 2, NA)))
  }
}
```


##1.2 Loci filtering
```{r}
## 1.2 Loci filtering

# Locus missingness
missing_per_locus <- rowMeans(is.na(geno_num))

# Minor allele count (MAC)
mac <- apply(geno_num, 1, function(x) {
  x <- x[!is.na(x)]
  # count whichever allele is rarer
  min(sum(x), 2*length(x) - sum(x))
})

# Keep loci: missingness ≤ 20%, MAC ≥ 3 (adjust threshold as needed)
keep_loci <- which(missing_per_locus <= 0.2 & mac >= 3)

# Apply locus filter
vcf_filtered <- vcf[keep_loci, ]

# Individual filtering
gt_filtered <- extract.gt(vcf_filtered, element = "GT", as.numeric = FALSE)
missing_per_ind <- colMeans(is.na(gt_filtered))
keep_inds <- names(missing_per_ind[missing_per_ind <= 0.5])
vcf_filtered <- vcf_filtered[, c(TRUE, colnames(gt_filtered) %in% keep_inds)]

# Remove 2020 individuals
keep_inds <- colnames(vcf_filtered)[-1]
vcf_filtered <- vcf_filtered[, c(TRUE, !grepl("-20", keep_inds))]
```

#Merge with Metadata
```{r}
pdam_gen <- vcfR2genind(vcf_filtered)

# Read metadata
meta <- fread("merged_metadata.csv")

# Check for match
inds <- indNames(pdam_gen)
inds_clean <- trimws(inds)
meta_clean <- meta %>% mutate(SampleID = trimws(as.character(SampleID)))

# Filter out '-20' samples from inds
inds_filtered <- inds_clean[!grepl("-20|-24b", inds_clean)]


# Filter metadata to match and remove '-20' individuals
meta_ordered_pdam <- meta_clean %>%
  filter(SampleID %in% inds_filtered) %>%
  arrange(match(SampleID, inds_filtered))

# Also update genind object to remove those individuals
pdam_gen <- pdam_gen[!grepl("-20|-24b", indNames(pdam_gen)), ]


```


##Add population level assignments 
```{r}
# 3. Ensure metadata rows are ordered to match genind individuals
# indNames() gives the individual IDs in the genind
meta <- meta_ordered_pdam %>%
  filter(SampleID %in% indNames(pdam_gen)) %>%   # keep only samples in genind
  arrange(match(SampleID, indNames(pdam_gen)))   # reorder to match genind

# 4. Assign population/reef to genind
pop(pdam_gen) <- as.factor(meta$Reef)

# 5. Verify
table(pop(pdam_gen))  # should show number of individuals per reef
```

```{r}
pdam_gen_full <- pdam_gen
save(pdam_gen_full, file = "pdam_gen_full.RData")
save(meta_ordered_pdam, file = "meta_ordered_pdam.RData")
```


# Section 2. PCA

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(vcfR)
library(adegenet)
library(data.table)

load("pdam_gen_full.RData")
load("meta_ordered_pdam.RData")
```


```{r}
x <- tab(pdam_gen_full, NA.method = "mean")
pca1 <- dudi.pca(x, scannf = FALSE, scale = FALSE, nf = 3)

percent <- pca1$eig / sum(pca1$eig) * 100
barplot(percent, ylab = "Variance explained (%)", names.arg = round(percent, 1))
```

#Prepare PCA coordinates + add population info
```{r}
# Extract PCA coordinates
ind_coords <- as.data.frame(pca1$li)

# Add individual names
ind_coords$ind <- rownames(ind_coords)

# Add population info directly from genind object
ind_coords$pop <- pdam_gen_full@pop  # already filtered to match individuals

ind_coords <- ind_coords %>% rename(SampleID = ind)

ind_coords <- ind_coords %>% left_join(meta_clean, by = "SampleID")
```


```{r}
# Custom hex colors for the first 12 reefs
my_colors <- c(
  "#d85874", "#ce4f35", "#c78048", "#6db543", "#3e8048",
  "#63be84", "#43bdc4", "#6486c9", "#7862cc", "#cb87c9",
  "#c34cae", "#9e496b"
)

# Reef order (14 reefs total)
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
                "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
                "Wreck", "Cato", "Elizabeth", "Middleton")

# Assign the custom colors to the first 12 reefs
custom_palette <- setNames(my_colors, reef_order[1:12])

# Assign custom colors to Elizabeth and Middleton
custom_palette["Elizabeth"] <- "#c4af46"
custom_palette["Middleton"] <- "#7a7932"

# Final named palette
reef_palette <- custom_palette

# View or use reef_palette
reef_palette
```

##2.6 Coral Sea PCA - reef
```{r}
library(ggplot2)
reef_order <- c("Willis", "Madgelaine", "Herald", "Diamond", "Coringa",
  "Marion", "Mellish", "Frederick", "Saumarez", "Kenn",
  "Wreck", "Cato", "GBR", "Elizabeth", "Middleton")

ind_coords$Reef <- factor(ind_coords$Reef, levels = reef_order)

ind_coords <- ind_coords %>% filter(Reef != "NA")

save(ind_coords, file = "ind_coords.RData")

pca.dam <- ggplot(ind_coords, aes(x = Axis1, y = Axis2, color = Reef)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "Reef"
  ) +
  scale_color_manual(values = reef_palette) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )

pca.dam
```

#PCA with WGS samples highlighted

```{r}
wgs_samples <- c(
"A025-24", "A035-24", "A235-24","A052-24", "A196-24", "A066-24", "A214-24", "A334-24", "A342-24", "A353-24", "A100-24", "A115-24", "A277-24", "A290-24",
"I015-24", "I034-24", "I066-24", "I205-24", "I209-24", "I113-24", "I095-24", "I171-24", "I214-24", "I053-24", "I108-24", "I126-24",
"M124-24", "M132-24", "M229-24", "M047-24", "M228-24", "M128-24", "M027-24", "M069-24", "M102-24", "M107-24", "M111-24", "M224-24", "M186-24","M247-24", "M263-24",
"PD046-24", "PD036-24", "PD141-24", "PV119-24", "PV167-24","PV286-24", "PV314-24", "PM124-24", "PM384-24", "PM471-24"
)

library(dplyr)

ind_coords <- ind_coords %>%
  mutate(WGS = if_else(SampleID %in% wgs_samples, "yes", "no")) %>% 
  mutate(WGS = as.factor(WGS))
```


```{r}
ggplot(ind_coords, aes(x = Axis1, y = Axis2)) +
  geom_point(
    data = subset(ind_coords, WGS == "no"),
    aes(color = WGS),
    size = 3, alpha = 0.9
  ) +
  geom_point(
    data = subset(ind_coords, WGS == "yes"),
    aes(color = WGS),
    size = 3, alpha = 0.9
  ) +
  scale_color_manual(
    values = c("no" = "grey70", "yes" = "darkred")
  ) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)"),
    color = "WGS"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5)
  )
```

```{r}
#adding host cluster so we can add shape by host 
ind_coords <- ind_coords %>% mutate(HostCluster = case_when(
    Axis1 < 0  ~ "PDC1",
    Axis1 > 0 ~ "PDC2",
    TRUE ~ NA_character_  # anything that doesn't match the above
  ))

ind_coords_pdam <- ind_coords
```

```{r}
pdam_cluster <-  ggplot(ind_coords_pdam, aes(x = Axis1, y = Axis2, colour = HostCluster)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  theme_minimal()
pdam_cluster
```
```{r}
 ggsave(pdam_cluster,
        file = "pdam_PCA_cluster.PNG",
        height = 189,
        width = 189,
        unit = "mm")
```

## save Cluster ind_coords

```{r}
# Create vectors of SampleIDs for each species based on pdamID
inds_PDC1 <- ind_coords_pdam$SampleID[ind_coords_pdam$HostCluster == "PDC1"]
inds_PDC2 <- ind_coords_pdam$SampleID[ind_coords_pdam$HostCluster == "PDC2"]


# Subset the genind object for each species
pdam_gen_PDC1 <- pdam_gen_full[inds_PDC1, ]
pdam_gen_PDC2 <- pdam_gen_full[inds_PDC2, ]



save(pdam_gen_PDC1, file = "pdam_gen_PDC1.RData")
save(pdam_gen_PDC2, file = "pdam_gen_PDC2.RData")
```

```{r}
# PCA for Lord Howe
x_PDC1 <- tab(pdam_gen_PDC1 , NA.method = "mean") #issue that PCA requires non-zero variance, and suggests that loci are monomorphic.

#for PCA to work the var cannot be 0
var_cols <- apply(x_PDC1, 2, var, na.rm = TRUE)
sum(var_cols == 0, na.rm=TRUE)  # monomorphic loci


# remove loci with zero variance (to get PCA to work)
x_PDC1 <- x_PDC1[, !is.na(var_cols) & var_cols > 0]

pca_PDC1 <- dudi.pca(x_PDC1, scannf = FALSE, scale = FALSE, nf = 3)
# # check for remaining NA 
# sum(is.na(x_PDC1))

# PCA for Coral Sea 
x_PDC2 <- tab(pdam_gen_PDC2, NA.method = "mean")
pca_PDC2 <- dudi.pca(x_PDC2, scannf = FALSE, scale = FALSE, nf = 3)

```

```{r}
# For PDC1
ind_coords_PDC1 <- as.data.frame(pca_PDC1$li)
ind_coords_PDC1$SampleID <- rownames(ind_coords_PDC1)
ind_coords_PDC1 <- left_join(ind_coords_PDC1, meta_ordered_pdam, by = "SampleID")
save(ind_coords_PDC1, file = "ind_coords_PDC1.RData")
# For PDC2
ind_coords_PDC2 <- as.data.frame(pca_PDC2$li)
ind_coords_PDC2$SampleID <- rownames(ind_coords_PDC2)
ind_coords_PDC2 <- left_join(ind_coords_PDC2, meta_ordered_pdam, by = "SampleID")
save(ind_coords_PDC2, file = "ind_coords_PDC2.RData")


```

```{r}
library(plotly)

# add tooltip aesthetics
pca.symbiont <- ggplot(ind_coords, 
                       aes(x = Axis1, y = Axis2, 
                           fill = Reef, 
                           shape = HostCluster,
                           text = paste0(
                             "Sample: ", SampleID, "<br>",
                             "Population: ", Reef, "<br>",
                             "Host Cluster: ", HostCluster
                           ))) +
  geom_point(size = 3, alpha = 0.9, color = "black") +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.3) +
  labs(
    x = paste0("PC1 (", round(pca1$eig[1] / sum(pca1$eig) * 100, 1), "%)"),
    y = paste0("PC2 (", round(pca1$eig[2] / sum(pca1$eig) * 100, 1), "%)")
  ) +
  scale_fill_manual(values = reef_palette) +
  scale_shape_manual(values = c(21, 22, 23, 24, 25)) +
  guides(
    fill = guide_legend(override.aes = list(shape = 21, color = "black")),
    shape = guide_legend(override.aes = list(fill = "grey70"))
  ) +
  theme_minimal()

# make interactive with custom tooltip
pca.symbiont.interactive <- ggplotly(pca.symbiont, tooltip = "text")

pca.symbiont.interactive
```

```{r}
#How many total individuals 
#How many per reef
# Total count
total_individuals <- ind_coords %>%
  summarise(n_total = n())

print(total_individuals)

# Count per reef
individuals_per_reef <- ind_coords %>%
  group_by(Reef) %>%
  summarise(n_individuals = n()) 

print(individuals_per_reef)
```

# Section 3. Genetic Diversity Stats
#Load Libraries
```{r}
# packages you'll need
library(tidyverse)
library(vroom)      # fast reading
library(vcfR)       # if working with VCF directly
library(adegenet)   # PCA on genotypes (if using genind)
library(lfmm)       # or lfmm2 (install from CRAN/Bioconductor)
library(vegan)      # RDA / partial RDA
library(car)        # vif
library(qqman)      # manhattan
library(hierfstat)
library(dplyr)
```

Population statistics 

#Load data
```{r}
load("pdam_gen_full.RData")
load("pdam_gen_PDC1.RData")
load("pdam_gen_PDC2.RData")
pdam_gen_PDC1
pdam_gen_PDC2


# Keep only individuals that do NOT contain '24b' (case-insensitive just in case)
pdam_gen_full <- pdam_gen_full[!grepl("24b", indNames(pdam_gen), ignore.case = TRUE), ]

# Keep only the part up to "-24"
clean_names <- sub("(-24).*", "\\1", indNames(pdam_gen_full))
indNames(pdam_gen_full) <- clean_names
```

```{r}
load("meta_ordered_pdam.RData")

meta_ordered_pdam <- meta_ordered_pdam %>%  
  mutate(MarinePark = case_when(
    Reef %in% c("Elizabeth", "Middleton") ~ "LH",
    !Reef %in% c("Elizabeth", "Middleton") ~ "CS"
  ))
```

# PDC1
## 1.1 Heterozygosity (Ho)
```{r}
pop_vector <- meta_ordered_pdam$MarinePark #extract pop info 
pop_vector <- pop_vector[1:nInd(pdam_gen_PDC1)]  #fixing mismatch issue between dataframes 
pop(pdam_gen_PDC1) <- as.factor(pop_vector) #assigns vector as pop info into the genind object 


#Here we are calculating basic stats using hierfstat
basic_pdam <- basic.stats(pdam_gen_PDC1, diploid = TRUE)

# Mean observed heterozygosity per site
Ho_pdam = apply(basic_pdam$Ho, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
Ho_pdam
```

## 1.2 Heterozygosity (He)
```{r}
# Mean expected heterozygosity per site
He_pdam = apply(basic_pdam$Hs, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
He_pdam
```

#Mean Ho and He pdamss populations
```{r}
# Mean Ho and He per population
mean_Ho <- apply(basic_pdam$Ho, 2, mean, na.rm = TRUE)
mean_He <- apply(basic_pdam$Hs, 2, mean, na.rm = TRUE)

round(cbind(mean_Ho, mean_He), 3)
```


##1.3 Inbreeding 
```{r}
apply(basic_pdam$Fis, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
```

```{r}
library(ggplot2)
library(tidyr)

fis_df <- as.data.frame(basic_pdam$Fis)
fis_long <- pivot_longer(fis_df, cols = everything(), names_to = "Population", values_to = "Fis")

ggplot(fis_long, aes(x = Fis)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  facet_wrap(~ Population, scales = "free_y") +
  theme_minimal() +
  labs(title = "Distribution of per-locus FIS per population")
```

# PDC2
## 1.1 Heterozygosity (Ho)
```{r}

pop_vector <- meta_ordered_pdam$MarinePark #extract pop info 
pop_vector <- pop_vector[1:nInd(pdam_gen_PDC2)]  #fixing mismatch issue between dataframes 
pop(pdam_gen_PDC2) <- as.factor(pop_vector) #assigns vector as pop info into the genind object 


#Here we are calculating basic stats using hierfstat
basic_pdam <- basic.stats(pdam_gen_PDC2, diploid = TRUE)

# Mean observed heterozygosity per site
Ho_pdam = apply(basic_pdam$Ho, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
Ho_pdam
```

## 1.2 Heterozygosity (He)
```{r}
# Mean expected heterozygosity per site
He_pdam = apply(basic_pdam$Hs, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
He_pdam
```

#Mean Ho and He pdamss populations
```{r}
# Mean Ho and He per population
mean_Ho <- apply(basic_pdam$Ho, 2, mean, na.rm = TRUE)
mean_He <- apply(basic_pdam$Hs, 2, mean, na.rm = TRUE)

round(cbind(mean_Ho, mean_He), 3)
```


##1.3 Inbreeding 
```{r}
apply(basic_pdam$Fis, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
```

```{r}
library(ggplot2)
library(tidyr)

fis_df <- as.data.frame(basic_pdam$Fis)
fis_long <- pivot_longer(fis_df, cols = everything(), names_to = "Population", values_to = "Fis")

ggplot(fis_long, aes(x = Fis)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  facet_wrap(~ Population, scales = "free_y") +
  theme_minimal() +
  labs(title = "Distribution of per-locus FIS per population")
```
#Add population info to 
```{r}
pop_vector <- meta_clean$Reef #extract pop info 
pop_vector <- pop_vector[1:nInd(pdam_gen)]  #fixing mismatch issue between dataframes 
pop(pdam_gen) <- as.factor(pop_vector) #assigns vector as pop info into the genind object 

pdam_gen <- pdam_gen[!is.na(pop(pdam_gen)), ]
```


## 2.1 Heterozygosity (Ho)
```{r}
#Here we are calculating basic stats using hierfstat
basic_pdam = basic.stats(pdam_gen, diploid = TRUE)

# Mean observed heterozygosity per site
Ho_pdam = apply(basic_pdam$Ho, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
Ho_pdam


```

## 2.2 Heterozygosity (He)
```{r}
# Mean expected heterozygosity per site
He_pdam = apply(basic_pdam$Hs, MARGIN = 2, FUN = mean, na.rm = TRUE) %>%
  round(digits = 3)
He_pdam

```


```{r}
# Create a data.frame of site names, Ho and He and then convert to long format
Het_pdam_df = data.frame(Site = names(Ho_pdam), Ho = Ho_pdam, He = He_pdam) %>%
  melt(id.vars = "Site")

# Custom theme for ggplot2
custom_theme = theme(
  axis.text.x = element_text(size = 10, angle = 90, vjust = 0.5, face = "bold"),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 12),
  axis.title.x = element_blank(),
  axis.line.y = element_line(size = 0.5),
  legend.title = element_blank(),
  legend.text = element_text(size = 12),
  panel.grid = element_blank(),
  panel.background = element_blank(),
  plot.title = element_text(hjust = 0.5, size = 15, face="bold")
  )

# Italic label
hetlab.o = expression(italic("H")[o])
hetlab.e = expression(italic("H")[e])

# pdampora heterozygosity barplot
ggplot(data = Het_pdam_df, aes(x = Site, y = value, fill = variable))+
  geom_bar(stat = "identity", position = position_dodge(width = 0.6), colour = "black")+
  scale_y_continuous(expand = c(0,0), limits = c(0,0.8), breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8))+
  scale_fill_manual(values = c("pink", "#bdbdbd"), labels = c(hetlab.o, hetlab.e))+
  ylab("Heterozygosity")+
  ggtitle("Pocillopora Damicornis")+
  custom_theme
```


##2.6 Calculate Fst
```{r}
#using hierfstat package here. 
#compute pairwise Fst values. Returns pairwise distances following Weir & Cockerham 1984 
pdam_fst = genet.dist(pdam_gen, method = "WC84") %>% round(digits = 3)
pdam_fst

save(pdam_fst, file = "pdam_fst.RData")

  #Elizabeth Middleton Frederick  Cato Wreck Marion  Kenn Mellish Flinders Moore Saumarez Willis
#Middleton      0.005                                                                                    
#Frederick      0.252     0.267                                                                          
#Cato           0.285     0.298     0.025                                                                
#Wreck          0.273     0.286     0.020 0.003                                                          
#Marion         0.279     0.290     0.029 0.034 0.033                                                    
#Kenn           0.269     0.283     0.005 0.020 0.012  0.037                                             
#Mellish        0.283     0.293     0.032 0.046 0.045  0.040 0.046                                       
#Flinders       0.298     0.311     0.020 0.045 0.034  0.029 0.033   0.027                               
#Moore          0.300     0.313     0.015 0.054 0.035  0.045 0.020   0.058    0.035                      
#Saumarez       0.145     0.162     0.038 0.063 0.060  0.063 0.047   0.068    0.049 0.040                
#Willis         0.292     0.305     0.026 0.040 0.038  0.030 0.045   0.015    0.023 0.063    0.067       
#Madgelaine     0.289     0.302     0.024 0.036 0.034  0.025 0.039   0.014    0.015 0.057    0.063  0.002
#Diamond        0.280     0.291     0.021 0.028 0.028  0.010 0.035   0.020    0.024 0.051    0.061  0.011
#Coringa        0.288     0.299     0.030 0.040 0.040  0.031 0.049   0.014    0.027 0.069    0.072  0.003
#Herald         0.291     0.304     0.012 0.030 0.024  0.016 0.028   0.016    0.010 0.056    0.045  0.009
#           Madgelaine Diamond Coringa
                         
#Diamond         0.008                
#Coringa         0.003   0.009        
#Herald          0.002   0.006   0.008
```


##2.7 Visualise Fst
```{r}

#Visualise pairwise Fst for reefs 
# Desired order of labels

lab_order = c("Elizabeth", "Middleton", "Wreck", "Kenn", "Mellish", "Marion", "Frederick")


# Change order of rows and cols
fst.mat = as.matrix(pdam_fst)
fst.mat1 = fst.mat[lab_order, ]
fst.mat2 = fst.mat1[, lab_order]

# Create a data.frame
ind = which(upper.tri(fst.mat2), arr.ind = TRUE)
fst.df = data.frame(Site1 = dimnames(fst.mat2)[[2]][ind[,2]],
                    Site2 = dimnames(fst.mat2)[[1]][ind[,1]],
                    Fst = fst.mat2[ ind ])

# Keep the order of the levels in the data.frame for plotting 
fst.df$Site1 = factor(fst.df$Site1, levels = unique(fst.df$Site1))
fst.df$Site2 = factor(fst.df$Site2, levels = unique(fst.df$Site2))
fst.df$Fst[fst.df$Fst < 0] = 0
fst.df %>% str
fst.label = expression(italic("F")[ST])
mid = max(fst.df$Fst) / 2

# Load required package
library(RColorBrewer)

# Plot heatmap with Spectral palette
ggplot(data = fst.df, aes(x = Site1, y = Site2, fill = Fst)) +
  geom_tile(colour = "black") +
  scale_fill_gradientn(
    colours = rev(brewer.pal(11, "Spectral")),  # Spectral palette reversed for better heatmap logic
    limits = c(0, 0.35),
    breaks = c(0, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3, 0.35),
    guide = "colorbar",
    name = fst.label
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0), position = "right") +
  theme(
    axis.text = element_text(colour = "black", size = 10, face = "bold"),
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 0.5),
    panel.grid = element_blank(),
    panel.background = element_blank(),
    legend.position = c(0.15, 0.7),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 10),
    legend.direction = "vertical"
  ) +
  coord_fixed(ratio = 0.9)
```



#3 Lord Howe IBD
##3.1 Data wrangle

```{r}
## --- Population-level IBD analysis ---

library(adegenet)
library(ade4)
library(MASS)      # for kde2d
library(scales)    # for alpha transparency

# 1. Convert genind → genpop (collapses individuals by pop)
genpop_obj <- genind2genpop(pdam_gen)

# 2. Genetic distance (Nei’s distance, method=2)
Dgen <- dist.genpop(genpop_obj, method = 2)

# 3. Geographic distance
reef_coords <- read.csv("pdam.coord.csv", row.names = 1)

# 4. Subset reef_coords to match the populations in genpop_obj
coords_cs <- reef_coords[popNames(genpop_obj), ]

# 5. Check alignment
all(rownames(coords_cs) == popNames(genpop_obj))
```


```{r}

# Euclidean distance among reefs
Dgeo <- dist(coords_cs)

# 4. Mantel test
ibd <- mantel.randtest(Dgen, Dgeo, nrepet = 999)
plot(ibd)

# 5. Pretty scatter plot with density overlay
Dgen_vec <- as.vector(Dgen)
Dgeo_vec <- as.vector(Dgeo)

dens <- kde2d(Dgeo_vec, Dgen_vec, n = 300)
myPal <- colorRampPalette(c("white","blue","gold","orange","red"))

plot(Dgeo_vec, Dgen_vec, pch = 20, cex = 0.8,
     xlab = "Geographic distance", ylab = "Genetic distance")
image(dens, col = alpha(myPal(300), 0.6), add = TRUE)
abline(lm(Dgen_vec ~ Dgeo_vec), lwd = 2)
lines(loess.smooth(Dgeo_vec, Dgen_vec), col = "red", lwd = 2)
```
**We see stronger trends of IBD in P. damicornis compared to A. humilis due to the brooding nature of this coral and therefore lower gene flow potential between reefs** 

```{r}
ibd_test <- mantel.randtest(Dgen, Dgeo, nrepet = 999)  # 999 permutations

ibd_test

IBD <- vegan::mantel(Dgen,log(Dgeo), method="pearson")
IBD

IBD <- vegan::mantel(Dgen,log(Dgeo), method="spearman")
IBD
```

#Log geographic distance vs. genetic distance 
```{r}
par(mar=c(4,4,0,0))
dens <- MASS::kde2d(log(Dgeo), Dgen, n=300)
plot(log(Dgeo), Dgen, pch=20, cex=0.5,  
    xlab="Log of Geographic Distance", ylab="Genetic Distance")
image(dens, col=transp(myPal(300), 0.7), add=TRUE)
abline(lm(Dgen ~ log(Dgeo)))
lines(loess.smooth(log(Dgeo), Dgen), col="red")
```

# Section 4. Gene environment analysis
